---
title: "Module 2 Assignment 3 tidyverse"
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Curious how we would do all of this in `tidyverse`? 

It takes fewer steps, but we haven't been using `tidyverse` very often, and not in this way before.

```{r message = FALSE}
library(tidyverse)
grasshoppers <- read_csv("grasshoppers.csv")
```


### Step 1: Calculate Population Parameter Estimates per Stratum

Use `group_by()`, `summarize()`, and `mutate()` to calculate parameter estimates
```{r}
grasshoppers <- grasshoppers %>% 
  # create groups for each stratum
  group_by(stratum) %>% 
  # calculate the means, standard deviations, and variances for each stratum
  summarise(y_bar = mean(abund),
            sd = sd(abund),
            var = var(abund),
            n = n()) %>% # how many rows per stratum AKA how many units per stratum
  # create a new column to put in the total number of units per stratum
  mutate(N = if_else(stratum == "A", 200,
                     if_else(stratum == "B", 500, 300)),
         tau_hat = y_bar * N) # calculate totals per stratum
grasshoppers
```

### Step 2: Calculate Measures of Uncertainty per Stratum

Use `mutate()` to calculate variance of the estimates for each stratum.
```{r}
grasshoppers <- grasshoppers %>% 
  # create new columns for:
  # 1) finite population correction factor, 
  # 2) variance of the estimate of the mean, and
  # 3) variance of the estimate of the total
  mutate(pop_correction = (N-n)/N,
         var_y_bar = pop_correction * (var/n),
         var_tau_hat = N^2 * var_y_bar)
grasshoppers
```

### Step 3: Calculate Overall Totals and Means

Use the `summarize()` function again to calculate the overall values.
```{r}
grasshoppers_overall <- grasshoppers %>% 
  summarize(overall_tau_hat = sum(tau_hat),
            overall_var_tau_hat = sum(var_tau_hat),
            overall_y_bar = overall_tau_hat/sum(N),
            overall_var_y_bar = overall_var_tau_hat / sum(N)^2) %>% 
  slice(1) # still thinks we want groups, so just select the first row (all rows have the same values)
grasshoppers_overall
```

