---
title: "Equation Sheet for Final Exam"
output: github_document
date: "2024-12-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Originally created by Jordan Stapley (12/11/2024)
Modified by Ellen Bledsoe (12/16/2024)

# Simple Random Sample

Index:

n = Number of sampled units N = total number of units sampled y_bar = sample mean s\^2 = sample varience mu (µ) = population mean sigma = population variance s = std dev SE = std err s(y_bar) = std error of the mean tau = population total tau_hat = estimate parameter T = population total

```{r}

data <- c(1,2,3,4)

##Sample mean (y_bar)
y_bar <- mean(data)

##Sample Variance (s^2)
s_2 <- var(data)

#Standard Std. Dev. (s)
s <- sd(data)

#Variance of estimate of the mean (var_y_bar)
N <- 
n <- 
k <- N/n

#If N is known 

var_y_bar <- ((N-n)/N)*(s_2/n)

#If N is unknown 

var_y_bar <- (s_2/n)  

#Std error of mean

#N is known
se_y_bar <-  sqrt(var_y_bar)

#N is unknown
se_y_bar <- sqrt(var_y_bar)

#Finite poulation correction 
((N-n)/N)

#Estimate of population total (tau_hat)

tau_hat = N*y_bar

#Variance of the estimate population total (var_tau_hat)

var_tau_hat <- (N^2) * (var_y_bar) 

#std error of population total

se_tau <- sqrt(var_tau_hat)

```

## Confidence Intervals

```{r}
#Alpha value 

alpha95 <- 0.05
alpha90 <- 0.10

#degrees of freedom 

df <- (n-1)

#t-value

t95 <- qt(1-(alpha95)/2, df)
t90 <- qt(1-(alpha90)/2, df)

#half width 

halfwidth_95 <- (t95*se_y_bar)
halfwidth_90 <- (t90*se_y_bar)

#Upper and lower CI

upperCI_95 <- y_bar + halfwidth_95
lowerCI_95 <- y_bar - halfwidth_95
upperCI_95
lowerCI_95

#CI for an estimate of population total (tau_hat)
halfwidth_95 <- (t95*se_tau_hat)
halfwidth_90 <- (t90*se_tau_hat)

upperCI_95 <- tau_hat + halfwidth_95
lowerCI_95 <- tau_hat - halfwidth_95
upperCI_95
lowerCI_95
```

## Stratified Random Sampling

Index:

-   L = Number of strata
-   N_i = Number of sample units in stratum i

```{r}
# Total Number of sample units in sampling frame
N <- 
  
#Total number of sample units in sample
n <- 

#Sampling Units per stratum
N_A <- 
N_B <- 
N_C <- 
  
#Std Dev for each stratum 
sd_A <- 
sd_B <- 
sd_C <- 
  
#Cost/time
cost_A <- 
cost_B <- 
cost_C <- 


#Optimal Allocation**
  
opt_allo_A <- ((N_A*sd_A)/sqrt(cost_A))
opt_allo_B <- ((N_B*sd_B)/sqrt(cost_B))
opt_allo_C <- ((N_C*sd_C)/sqrt(cost_C))

# this line of code prints out all of the values   
opt_allo_A; opt_allo_B; opt_allo_C
  
opt_allo_total <- opt_allo_A+opt_allo_B+opt_allo_C
opt_allo_total
  
#Optimal units to sample

n_A <- round(n*(opt_allo_A / (opt_allo_total)), digits = 0)
n_B <- round(n*(opt_allo_B / (opt_allo_total)), digits = 0)
n_C <- round(n*(opt_allo_C / (opt_allo_total)), digits = 0)
  
n_A; n_B; n_C

```

### Estimates per Stratum

```{r}
# Estimates per Stratum

mean_A <- mean(stratumA$abund)
mean_B <- mean(stratumB$abund)
mean_C <- mean(stratumC$abund)
  
mean_A; mean_B; mean_C


tau_hat_A <- N_A * mean_A
tau_hat_B <- N_B * mean_B
tau_hat_C <- N_C * mean_C
  
tau_hat_A; tau_hat_B; tau_hat_C

var_A <-var(stratumA$abund)
var_B <-var(stratumB$abund)
var_C <-var(stratumC$abund)
  
var_A; var_B; var_C

```

### Uncertainity per Stratum

```{r}
pop_correction_A <- (N_A - n_A)/N_A
pop_correction_B <- (N_B - n_B)/N_B
pop_correction_C <- (N_C - n_C)/N_C

var_y_bar_A <- (pop_correction_A)*(var_A/n_A)
var_y_bar_B <- pop_correction_B*(var_B/n_B)
var_y_bar_C <- pop_correction_C*(var_C/n_C)
  
var_y_bar_A; var_y_bar_B; var_y_bar_C
```

### Variance of the Estimate of the Total (var_tau_hat)

```{r}
var_tau_hat_A <- N_A^2 * var_y_bar_A
var_tau_hat_B <- N_B^2 * var_y_bar_B
var_tau_hat_C <- N_C^2 * var_y_bar_C
  
var_tau_hat_A; var_tau_hat_B; var_tau_hat_C
```

## Overall Calculations: Stratified Sampling

Total Abundance

```{r}
tau_hat <- tau_hat_A + tau_hat_B + tau_hat_C
tau_hat
```

Variance of total abundance estimate

```{r}
var_tau_hat <- var_tau_hat_A + var_tau_hat_B + var_tau_hat_C
var_tau_hat
```

Overall Mean

```{r}
## Overall Population mean 

y_bar <- tau_hat/(N)
y_bar
```

Overall Variance of Estimate of the Mean

```{r}
var_y_bar <- var_tau_hat/(N^2)
var_y_bar

```


# Imperfect Sampling

C = Count p = detection probabilty tau = true abudnce

tau_hat = estimated abundance p_hat = estimated detection probability

```{r}
# Canonical Abundance Estimator

C <- 
p_hat <- 
tau <- 
alpha <- #n/N

tau_hat <- C/p_hat
tau_hat

```

## Occupancy

Units: S = number of sites in hte sample S_D = Number of sites where species was detected at least once K = Number of surveys per site psi = occupancy pi_j = probability that the species is detected at site i during survey j if the species is truly present

## Abundance

### Capture-Mark-Recapture

tau_hat = estimated abundance c1 = number of individuals of individuals captures and amrked in first sample c2 = Number of indiviuals captured in second sample (Marked + Unmarked) r2 = Number of indiviuals captured in second sample that were previously unmarked

```{r}
## Lincoln Peterson 
c1 <- 
c2 <- 
r2 <- 

tau_hat <- ((c1*c2)/r2)
tau_hat

det_prob <- r2 / c1
det_prob

```

### Constant-Effort Removal

```{r}
##Removal methods 
c1 <- 
c2 <- 
  
tau_hat <- (c1^2) / (c1-c2)
tau_hat
  
SE_t_hat <- ((c1*c2)*(sqrt(c1 + c2))) / ((c1-c2)^2)
SE_t_hat
```

### Catch per Unit Effort (CPUE)

```{r}
y <- 
x <- 
k <- 
tau_hat <- x + (y / -k)
tau_hat
```

## Density

```{r}
##Distance Sampling -- Density

C <- 
w <- 
l <- 

D_hat <- ((C)/(2*w*l))
D_hat
```

## Species Diversity

```{r}
# Shannon Index
data1 <- c(1,2,3,4 )
total <- sum(data1)
prop <- data1/total
ln_prop <-  log(prop)
lnprop_prop <- prop * ln_prop
H <- sum(lnprop_prop) * -1 
H

# Simpson Index
data1 <- c(1,2,3,4 )
total <- sum(data1)
prop <- data1/total
prop2 <-  prop^2
D <- 1 - sum(prop2)
D

```
