---
title: 'Module "4": Assignment 1'
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment Details

### Purpose

The goal of this assignment is to work with community data to calculate observed and estimated species richness and plot species accumulation curves.

### Task

Write R code to successfully answer each question below or write text to successfully answer the question.

### Criteria for Success

-   Code is within the provided code chunks
-   Code chunks run without errors
-   Code produces the correct result
    -   Code attempts will get half points
    -   Code that produces the correct answer will receive full points
-   Text answers correctly address the question asked

### Due Date

May 12 at midnight MST

------------------------------------------------------------------------

# Assignment Questions

All questions are worth 1 point unless otherwise specified.

We are going to compare two different communities in this assignment.

## Set-Up

1.  We are going to need to load the community ecology package, `vegan`.

```{r message = FALSE}

```

## Community 1

For our first community in this assignment, we are going to use a site x species matrix for trail cameras along the Santa Cruz River.

2.  Use the `read.csv()` function to bring in the SCR trail camera data (called `SCR_camera_data.csv`)

```{r}
scr <- 
```

3.  First, use the `head()` function to take a look at the dataset.

```{r}

```

4. How many sites were surveyed? How many species were seen?

*Sites*: 
*Species*: 

5. Use the `specaccum()` function to prep the data for the species accumulation curve.

```{r}
# A warning is fine!
scr_sac <- 
scr_sac
```

6.  Next, plot the species accumulation curve based on our observed data.

```{r}

```

7.  Use the `specpool()` function to calculate the estimated species richness.

```{r}
scr_richness <- 
scr_richness
```
8. Save the observed species richness and estimated species richness values as objects. You can use the `$` operator to reference specific named values from the `scr_richness` object.

```{r}
scr_observed <- 
scr_observed

scr_estimated <- 
scr_estimated
```

9. Based on the Chao estimator, how many species are present but we did not observe?
```{r}

```

10. Now, let's add our estimated species richness to our species accumulation curve. Remember to set the y-axis range appropriately and to run the `plot()` and `abline()` functions toegther.

```{r}

```


11. Interpret the plot you just made. What is the predicted asymptote value? How does it relate the our species accumulation curve? What does that mean?

*Answer*: 

## Community 2

12. Now, bring in the data for our second community, fish species from the Doubs River in France.

```{r}
fish <- 
```

13.  Use the `head()` function to take a look at the dataset.

```{r}

```
In this dataset, each row represents a survey occasion rather than a site.

14. Use the `specaccum()` function to prep the data for the species accumulation curve.

```{r}
# A warning is fine!
fish_sac <- 
fish_sac
```

15.  Next, plot the species accumulation curve based on our observed data.

```{r}

```
16. Use the `specpool()` function to calculate the estimated species richness.

```{r}
fish_richness <- 
fish_richness
```

17. Save the observed species richness and estimated species richness values as objects. You can use the `$` operator to reference specific named values from the `scr_richness` object.

```{r}
fish_observed <- 
fish_observed

fish_estimated <-
fish_estimated
```

18. Based on the Chao estimator, how many species are present but we did not observe?
```{r}

```

19. Now, let's add our estimated species richness to our species accumulation curve. Remember to set the y-axis range appropriately and to run the `plot()` and `abline()` functions together.

```{r}

```

20. Interpret the plot you just made. How does community 2 differ from community 1? How does our sampling differ between community 1 and community 2? 

*Answer*: