---
title: "Module 2 Assignment 1"
author: ""
date: '2022-10-11'
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Simple Random Sampling -- Rocks!

## Assignment Details

### Purpose

The goal of this assignment is to use R to calculate and understand
accuracy of estimates (standard error specifically).

### Task

Write R code to successfully answer each question below or write text to
successfully answer the question.

### Criteria for Success

-   Code is within the provided code chunks
-   Code chunks run without errors
-   Code produces the correct result
    -   Code attempts will get half points
    -   Code that produces the correct answer will receive full points
-   Text answers correctly address the question asked

### Due Date

March 16 by 2 pm MST

------------------------------------------------------------------------

# Assignment Questions

In this assignment, you will be using the data we collected during the
last lab to calculate (a) estimates of population parameters and (b) the
amount of uncertainty in those estimates.

Remember, each group used a simple random sampling design to sample a
"population" of dark rocks; your large square represents the sampling
frame, while each smaller square (\~4 in x 4 in) represents a sample
unit.

## Short Answer

Answer the following question with respect to the data we collected last
week (AKA, using phrases like "large square," "little square," "dark
rocks," etc.)

Think back to how we sampled the rock data. Answer the following
question in terms of our data (aka, "big square," "little squares,"
"dark rocks," etc.)

1.  Define the *sample* standard deviation in terms of our data (not the
    equation but what we use to calculate standard deviation, what it
    tells us, and what it represents).

    *Answer:*

2.  Define the standard error of the estimate of the mean (again, in
    terms of our data and what it represents).

    *Answer:*

3.  How do the *sample standard deviation* and *standard error of the
    estimate of the mean* differ? (If you've explained that above,
    reiterate it here.)

    *Answer:*

4.  What would happen to the *sample standard deviation* if we sampled
    20 sampling units (little squares) instead of only 8? Would it
    increase, decrease, or stay about the same? How about the *standard
    error*? Explain your rationale for both.

    *Answer:*

### Set-up:

If you haven't already saved and shared your data from last week with
your group:

(a) Get your (or your group's) simple random sampling rock data.
(b) Make sure it has been saved as a .csv file. If it hasn't, you will
    need to do so using "Save As".
(c) In Posit Cloud, using the "Upload" button in the lower right-hand
    panel, upload your .csv file.
(d) In the Files tab in the bottom right-hand corner, check the box next
    to the .csv file that you just uploaded, then click the "Rename"
    button (near "Upload").
(e) If needed, rename your .csv file as "rocks_abund.csv"

### Important Notes:

(a) I recommend referencing the slides from class! Remember, equations
    don't show on D2L, so you'll want to download the slides or use the
    PDF version.
(b) I also recommend taking a look back at Module 1 Assignment 3 to
    remind yourself how to work with data frames.
(c) Because each group will have different values for your number of
    rocks, the answer key won't directly match your values. I've
    included the example dataset that I used to create the answer key if
    you want to check your values.

## Calculations

Each question is worth 1 point unless otherwise noted.

5.  Use the `read.csv()` function from the tidyverse to read in your
    rocks file. Save this data in an object called `rocks`.

    Below, replace the file name inside the quotation amrks with the
    file name of your rock data (also in quotation marks).

```{r}
# when you read in your file, the name of the file needs to be inside quotation marks
# it also needs to have the .csv at the end, as shown below
rocks <- read.csv("your_file_name_here.csv")

# if your rocks data frame has more than 8 observations (check in the environment), run this line of code:
rocks <- na.omit(rocks)
```

6.  Create the object `N`, which is the number of units in your
    *sampling frame*. No functions needed; type the number and assign it
    to `N`.

    *Remember: after you have created a new object, type out the name of
    that object on the line of code below so that the value of that
    object will print out when you knit to PDF. This helps us out while
    grading quite a bit!*

```{r}

```

7.  Create the object `n`, which is your *sample size,* the number of
    sample units that make up your sample.

    To do this, use the `length()` function to count the number of
    elements in the `sample_unit` column of the `rocks` data frame.
    Remember, we can use the `$` to reference a specific column in a
    data frame.

    The value might have an L at the end of it for "length," but it will
    act like a number in subsequent code.

```{r}

```

### Estimating the Mean

One way we can calculate values is to call the column that we want and
then save that value to an object.

For example, let's say we wanted to calculate the median of our dark
stone abundance data. We could run the following line of code:
`rocks_median <- median(rocks$abund_dark_stones)`

For the remainder of the coding questions below, remember to print out
the answer by typing the name of the object again, below your line of
code.

In the above example, it would look like:

```{r, eval=FALSE}
rocks_median <- median(rocks$abund_dark_stones)
rocks_median
```

8.  Create an object called `rocks_mean` that contains the sample mean
    (AKA the mean number of dark stones per sample unit). Follow the
    code example above.

    *Again, after you have created a new object, type out the name of
    that object on the line of code below so that the value of that
    object will print out when you knit to PDF. Do this for the rest of
    the assignment!*

```{r}

```

9.  Create objects called `rocks_var` and `rocks_sd` which contain the
    sample variance and sample standard deviation, respectively.

    *Hint: use `var()` and `sd()` functions.*

```{r}

```

### Calculating Uncertainty for the Estimate of the Mean

We have calculated our estimates for multiple population parameters
(mean, standard deviation).

Now we are moving on to calculating a measure of uncertainty.

R doesn't have a function for calculating the standard error (SE) of the
mean, so let's calculate it ourselves using the equation from class.

10. First, do you remember that finite population correction factor?
    There's no function for it, so let's create an object called
    `pop_correction` using that formula.

    *Hint #1: use the objects we've already created (`N`, `n`; don't
    enter any numbers directly*

    *Hint #2: use parentheses to group the values in the numerator
    together*

```{r}

```

11. Why are we using the finite population correction factor in our
    formulas?

    *Answer:*

12. Using the objects you've already created and the formula (check the
    lecture slides from this module!), create an object called
    `rocks_var_ybar` which contains the variance of the *estimate* of
    the mean.

    Remember, this is different from the sample variance that you've
    already calculated...

    There is no function for this, so you'll need to use the equation.

    \_Hint: if you want to multiple something in R, you must use \*.\_

```{r}

```

13. Calculate the SE of the estimate of the mean by taking the square
    root (`sqrt()`) of the `rocks_var_ybar` value. Save this as an
    object `rocks_SE_ybar`.

```{r}

```

### Estimating the Population (and the Uncertainty of the Estimate)

We've calculated standard error (SE) for the estimate of the mean above.
Now, let's calculate them for the estimate of the population abundance
(tau hat)

14. First, let's calculate the total number of rocks in the population.
    Save the output as `rocks_total`. Use objects you've already
    created.

```{r}

```

15. Now calculate the variance of the estimate of the population total,
    saving it as `rocks_var_total`.

```{r}

```

16. Calculate the SE of the estimate of the population total, saving it
    as `rocks_SE_total`.

```{r}

```

17. Write a short paragraph (3-5 sentences) putting the values that
    we've just calculated into context for our dark rocks sampling
    adventure. (3 points)

    In your answer, be sure to address the following questions and use
    the correct units of measurement:

    -   According to your calculations, approximately how many dark
        rocks were there in your big square (AKA your dark rock
        "population")? What is your calculated uncertainty in your
        estimate?
    -   How much variation in the number of dark rocks per little square
        do you estimate there to be?
    -   What is your estimated average \# dark rocks per little square
        (AKA your dark rock density)? What is your calculated
        uncertainty in your estimate in the density?

*Answer*:
