---
title: "Module 2, Assigment 1"
author: "Ellen Bledsoe"
date: '2022-09-19'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment Details

### Purpose

### Task

### Criteria for Success

### Due Date

# Assignment Questions

## Definitions (2 point each)

*In your own words*, define/describe the following terms. These don't
need to be technical descriptions but rather how you are thinking about
them.

1.  *Reproducibility*:

2.  *Open science*:

3.  *R*:

4.  *RStudio*:

## Short-Answer Question (2 points)

5.  In 2-3 sentences, explain why data management is important for
    reproducibility.

*Answer:*

## Vectors (1 point each)

6.  Multiply 8 x 5.

```{r}

```

7.  Create the object `x` and have it contain the result of 40 divided
    by 2.77

```{r}

```

8.  Create the object `y` and have it contain the result of the square
    root of 9 (HINT: use the `sqrt` function).

```{r}

```

9.  Create the object `num_vec` that contains the values 1, 3, 5, 7

```{r}

```

10. Run the line below to create the object heights. Use the `mean()`
    function to calculate the mean value of height.

```{r}
heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63,
NA, 72, 65, 64, 70, 63, 65)
```

```{r}

```

11. This should yield an odd result caused by the `NA` values. To
    resolve this, use the help function to learn about the argument
    `na.rm = TRUE` that applies to many R functions

```{r}
help(mean) # same as ?mean
```

Issue a revised command to calculate the mean value in `heights` and
send the result to the console.

```{r}

```

12. Write and execute a line of code that selects the 6-10th height
    values.

```{r}
heights[6:10]
```

13. Execute the line below. What does this line of code do?

```{r}
heights67 <- heights[heights > 67]
```

*Answer:*

14. Execute the two lines below. What do these lines of code tell us?

```{r}
length(heights) 
length(heights67)
```

*Answer:*

15. Create a character vector called `rhymes` that contains the
    following values: cat, rat, bat, hat

```{r}

```

## Data Frames

Run the following line of code. Like the Palmer Penguins data set that
we used in class, `ratdat` is also a data package. It has data from my
PhD field site, the Portal Project. You can learn more about the Portal
Project [here](https://portal.weecology.org/).

We are going to be using the `surveys` data set from the `ratdat`
package.

```{r}
library(ratdat)
surveys
```

16. How many rows does this data frame have? How many columns?

    *Rows*:

    *Columns*:

17. Look at the first 6 rows of data. You can either do this by using a
    function or by using index subsetting (with the square brackets: [])

```{r}

```

18. Use a function to print the column names of surveys.

```{r}

```

19. Create a histogram with the hindfoot lengths column

```{r}

```

20. What does the following line of code do?

```{r}
surveys[surveys$species_id == "DM", ]
```

*Answer:*

## Calculating the Coefficient of Variance

The remainder of the questions in this assignment lead you through the
steps to calculate the coefficient of variance for two different species
at the field site: the largest (NL, the woodrat *Neotoma albigula*) and
one of the smallest (BA, the Northern pygmy mouse, *Baiomys taylori*).

21. Create 2 different data frames: one called `NL` with rows that have
    "NL" in the species_id column and one called `BA` with only rows
    which have "BA" in the species_id column. (Hint: take a look at
    question 20 for inspiration on how to do that)

```{r}
NL <- 
BA <- 
```

22. Calculate the mean weight for each species.

```{r}
mean_NL <- 
mean_BA <- 
```

23. In a similar fashion, calculate the standard deviation (`sd()`) for
    the weight of both species.

```{r}
sd_NL <- 
sd_BA <- 
```

24. Using the objects you just created in questions 22 and 23, calculate
    the coefficient of variance (CV) for each species.

```{r}
cv_NL <- 
cv_BA <- 
```

25. In 1-2 sentences, interpret the results of your calculations.

*Answer:*

## Assignment Submission Guidelines

When you are finished, I want you to export this .Rmd file and upload it
into the assignment dropbox on D2L.

To do you, follow these steps:

-   Make sure you have saved all of your code!
-   In the "Files" pane, check the empty box near this .Rmd file
-   Still in the "Files" pane, click the small downward arrow next to
    the gear ("More")
-   Select "Export"
-   Rename you file with the following pattern:
    "LastName_Module2_Assignment1.Rmd"
-   Once you've exported, the .Rmd file will be in your "Downloads"
    folder on your computer
-   Submit this file to D2L
