---
title: "Module 3 Assignment 1"
author: "Ellen Bledsoe"
date: '2022-10-11'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Simple Random Sampling -- Rocks!

## Assignment Details

### Purpose

The goal of this assignment is to use R to calculate and understand
accuracy of estimates.

### Task

Write R code to successfully answer each question below or write text to
successfully answer the question.

### Criteria for Success

-   Code is within the provided code chunks
-   Code is commented with brief descriptions of what the code does
-   Code chunks run without errors
-   Code produces the correct result
-   Text answers correctly address the question asked

### Due Date

Sept. 29 at midnight MST

------------------------------------------------------------------------

## Assignment Questions

### Short Answer (2 pt each)

1.  Define the sample standard deviation (not the equation but what
    stdev tells us).

2.  Define the standard error of the estimate of the mean (again, what
    is tells us).

3.  How do standard deviation and standard error differ? (If you've
    explained that above, reiterate it here.)

4.  What does a confidence interval tell us?

### Calculations

Each question is worth 1 point unless otherwise noted.

#### Set-up:

1.  Get your (or your group's) simple random sampling rock data.
2.  Make sure it has been saved as a .csv file. If it hasn't, you will
    need to do so using "Save As".
3.  In RStudio Cloud, using the "Upload" button in the lower right-hand
    panel, upload your .csv file.

Load the tidyverse. For this assignment, we will only be using this to
read in our .csv file.

```{r}
library(tidyverse)
```

5.  Use the `read_csv()` function from the tidyverse to read in your
    rocks file. Save this data in an object called `rocks`.

    Below, replace the file name with the file name of your rock data.

```{r}
rocks <- read_csv("your_file_name_here.csv") %>% drop_na()
```

6.  Create the object N, which is the number of units in your *sampling
    frame*.

```{r}
N <- 64
```

7.  Create the object n, which is your *sample size,* the number of
    sample units that make up your sample. To do this, use the `nrow()`
    function

    (Hint: if you don't remember how to use the `nrow()` function, run
    `?nrow`)

```{r}
n <- nrow(rocks)
```

### Estimating the Mean

One way we can calculate values is to call the column that we want and
then save that value to an object.

For example, let's say we wanted to calculate the median of our dark
stone abundance data. We could run the following line of code:
`rocks_median <- median(rocks$abund_dark_stones)`

Typically, I prefer using the `tidyverse` functions, but for performing
calculations like we are doing in this assignment, working with vectors
(as opposed to data frames) is easier; the code above does that for us.

8.  Create an object called `rocks_mean` that contains the sample mean
    (AKA the mean number of dark stones per sample unit). Follow the
    code example above.

```{r}
rocks_mean <- mean(rocks$abund_dark_stones)
```

9.  Create objects called `rocks_var` and `rocks_sd` which contain the
    sample variance and sample standard deviation, respectively. Hint:
    use `var()` and `sd()` functions.

```{r}
rocks_var <- var(rocks$abund_dark_stones)
rocks_sd <- sd(rocks$abund_dark_stones)
```

### Calculating Standard Error for the Mean

R doesn't have a function for calculating the standard error (SE), so
let's calculate it ourselves. From the equation

10. First, do you remember that finite population correction factor? We
    use that to modify SE when we know what N is. In this case, we do
    know N, so let's create an object called `pop_correction` using that
    formula: (N-n)/N

    (Hint: use the objects we've already created (`N`, `n`--don't enter
    any numbers directly)

```{r}
pop_correction <- (N-n)/N
```

11. Using the objects you've already created and the formula (check the
    lecture slides from this module!), create an object called
    `rocks_var_ybar` which contains the variance of the *estimate* of
    the mean.

    Remember, this is different from the sample variance that you've
    already calculated, but that sample variance is included...

```{r}
rocks_var_ybar <- pop_correction * (rocks_var/n)
```

12. Calculate the SE of the estimate of the mean by taking the square
    root of the `rocks_var_ybar` value. Call this object rocks_SE.

```{r}
rocks_SE <- sqrt(rocks_var_ybar)
```

### Calculating Confidence Intervals for the Mean

Hooray, we've now calculated the sample mean, sample variance, sample
standard deviation, and the standard error (among other things)! Let's
calculate a 95% confidence interval for our sample mean.

13. Create an object called `alpha` to represent our alpha value for a
    95% confidence interval. Because this is a constant, we can assign a
    number to it.

```{r}
alpha95 <- 0.05
```

14. Create an object called `df` for degrees of freedom. Use the object
    `n` in your calculation.

```{r}
df <- n - 1
```

We are going to use a function called `qt()` to find the t-value that we
need. Assuming you've created the objects `alpha` and `df` above,
running this code chunk will give us our t-value, or critical value, for
95% CI.

```{r}
t95 <- qt(1-(alpha95)/2, df)
```

15. Using `t95` and the standard error that we calculated earlier,
    create an object called `half_width` which contains the half-width
    value for the 95% CI.

```{r}
half_width <- t95 * rocks_SE
```

16. Create an object called `upperCI_mean` that contains the upper value
    of the CI for the estimate of the mean. Use objects you've already
    created to calculate it.

```{r}
upperCI_mean <- rocks_mean + half_width
```

17. Create an object called `lowerCI_mean` that contains the lower value
    of the CI for the estimate of the mean.

```{r}
lowerCI_mean <- rocks_mean - half_width
```

18. Report (AKA type out) the 95% CI for the estimate of the mean below.
    What does this range tell us about our data? (2 points)

### Estimating the Population (and Accuracy of the Estimate)

We've calculated standard error (SE) and confidence interval (CI) for
the estimate of the mean above. Now, let's calculate them for the
estimate of the population abundance (tau hat)

19. First, let's calculate the total number of rocks in the population.
    Call this `rocks_total`. Use objects you've already created.

```{r}
rocks_total <- N * rocks_mean
```

20. Now calculate the variance of the population total,
    `rocks_var_total`.

```{r}
rocks_var_total <- N * (N-n) * (rocks_var/n)
```

21. Calculate the SE of the estimate of the population total,
    `rocks_SE_total`.

```{r}
rocks_SE_total <- sqrt(rocks_var_total)
```

22. Calculate the half-width value for the population total estimate.

    (Hint: the values `alpha95` and `t95` don't change)

```{r}
half_width_total <- t95 * rocks_SE_total
```

23. Calculate and report the upper and lower 95% CI values for the total
    population of rocks.

```{r}
upperCI_total <- rocks_total + half_width_total
lowerCI_total <- rocks_total - half_width_total
```

24. What do these confidence intervals tell us about our data and our
    parameter?
