---
title: "Module 2, Assigment 1"
author: "Ellen Bledsoe" # <- your name here
date: '2022-09-19'      # <- current date here
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Purpose

The goal of this assignment is to explore computational reproducibility and apply the base R coding skills we've learned and practiced in class and lab.

### Task

Write R code to successfully answer each question below and/or write text to successfully answer the question.

### Criteria for Success

-   Code is within the provided code chunks
-   Code chunks run without errors
-   Code produces the correct result
    -   Code attempts will get half points
    -   Code that produces the correct answer will receive full points
-   Text answers correctly address the question asked

### Due Date

Feb 20 at midnight MST

# Assignment Questions

This assignment is worth 20 points total. Each question is worth 1 point.

## Definitions (1 point each)

*In your own words*, define/describe the following terms. These don't need to be technical descriptions but rather how you are thinking about them.

1.  *Reproducibility*:

2.  *Open science*:

3.  *R*:

4.  *RStudio*:

## Vectors (1 point each)

5.  Run the first code chunk below to create the object heights. Use the `mean()` function to calculate the mean value of height.

```{r}
heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63, NA, 72, 65, 64, 70, 63, 65)
```

```{r}

```

6.  This should yield an odd result caused by the `NA` values. To resolve this, use the help function to remind yourself about the argument `na.rm = TRUE` that applies to many R functions

```{r}
help(mean) # same as ?mean
```

Issue a revised command to calculate the mean value in `heights` and send the result to the console.

```{r}

```

7.  Write a line of code that selects the 6-10th height values.

```{r}

```

8.  Create a character vector called `rhymes` that contains the following values: cat, rat, bat, hat

```{r}

```

## Data Frames

**IMPORTANT!** Run the following code chunk to read in the cactus pad dataset. *You don't need to understand what is happening, though I've added some comments in case you are interested!*

We will be using that data for the remainder of the assignment.

Once you run the code chunk, the data frame will be saved in your environment as the object `pads`.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
# read in the cactus pads data
pads <- read_csv("../data_raw/CactusPadsData.csv") 

# read in the cactus data and choose only the group_id, species, and size columns
cactus <- read_csv("../data_raw/CactusData.csv") %>% select(group_id, species, size)

# join the two datasets together using the group_id column
# this matches up the cactus species and size for each cactus pad (row) with the same group_id
pads <- right_join(pads, cactus, by = "group_id")

# remove the cactus data frame because we don't need it anymore (and it might be confusing)
rm(cactus)
```

9.  How many rows does this data frame have? How many columns? You can either use code to figure this out or look at the object in the environment.

    *Rows*:

    *Columns*:

```{r}
# optional space for code to answer the question above

```

10. Look at the first 6 rows of data. You can either do this by using a function or by using index sub-setting.

```{r}

```

11. Calculate the mean of depth of all the cactus pads that were measured.

```{r}

```

12. Use the `round()` function to around the mean of the cactus pad depths (Q11) to only 1 place after the decimal.

```{r}

```

13. Create a histogram with the width of all the cactus pads that were measured.

```{r}

```

14. In 1-2 sentences, describe what each axis on this histogram represents.

*Answer:*

15. Describe what the following line of code is doing. Make sure to include each part of the code (i.e., before the arrow, the arrow, after the arrow...). (As a point of reference, this took me \~6 sentences in the answer key).

```{r}
pad_depth <- pads[pads$depth_in > 1, ]
```

*Answer:*

### Calculating the Coefficient of Variance

The remainder of the questions in this assignment lead you through the steps to calculate the coefficient of variance for pad length for two of the three cactus species that we measured. This will give us relative amounts of variation in between the two species.

16. Create 2 different data frames: one called `ficus` with only rows that have *Opuntia ficus-indica* in the species column and one called `santa_rita` with only rows which have *Opuntia santa-rita* in the species column. (Hint: take a look at Q15 for inspiration on how to do that)

```{r}
ficus <- 
santa_rita <- 
```

17. Calculate the mean pad length for each species.

```{r}
mean_ficus <- 
mean_santa <- 
```

18. In a similar fashion, calculate the standard deviation (`sd()`) for the weight of both species.

```{r}
sd_ficus <- 
sd_santa <- 
```

19. Using the objects (not just the numbers) you just created in questions 22 and 23, calculate the coefficient of variance (CV) for each species. Remember to multiply by 100 to convert into a percentage.

```{r}
cv_ficus <- 
cv_santa <- 
```

20. In 2-3 sentences, interpret the results of your calculations from Q17-19.

*Answer:*

# Turning in Your Assignment

Follow these steps to successfully turn in your assignment on D2L.

1.  Click the `Knit` button up near the top of this document. This should produce a PDF file that shows up in the `Files` panel on the bottom-right of your screen.
2.  Click the empty box to the left of the PDF file.
3.  Click on the blue gear near the top of the `Files` panel and choose Export.
4.  Put your last name at the front of the file name when prompted, then click the Download button. The PDF file of your assignment is now in your "Downloads" folder on your device.
5.  Head over to D2L and navigate to Module 1 Assignment 2. Submit the PDF file that you just downloaded.
