---
title: "Module 2 Assignment 2"
author: "Ellen Bledsoe"   # <- your name here
date: "`r Sys.Date()`"    # <- date here
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assessing Accuracy

## Assignment Details

### Purpose

The goal of this assignment is to use R to calculate and understand
accuracy of estimates through standard error and confidence intervals.

### Task

Write R code to successfully answer each question below or write text to
successfully answer the question.

### Criteria for Success

-   Code is within the provided code chunks
-   Code chunks run without errors
-   Code produces the correct result
    -   Code attempts will get half points
    -   Code that produces the correct answer will receive full points
-   Text answers correctly address the question asked

### Due Date

Oct 22 before lab

------------------------------------------------------------------------

# Assignment Questions

Each question is worth 1 point unless otherwise noted.

## Set-up:

For our assignment today, we will be using data from my PhD field site
in southeastern Arizona in the Chiricahuas. In fact, we will be using
some of the first data I ever collected for the project! To learn more
about the Portal Project, check out the project's
[website](https://portal.weecology.org/). Researchers have been
collecting data at this site monthly since 1977!

What you need to know for today is that this is small mammal (rodent)
data.

The site has 24 plots, although we will only be using 18 of them today
(6 of them are rodent removal plots, so they aren't very helpful in
estimating rodent densities!).

As such, `n` will be 18 for this assignment. We do not have a `N`.

The first thing we need to do is download the data. My PhD lab group
actually created our own R package for using the cleaned up data, so we
are going to load that package, called
[`portalr`](https://weecology.github.io/portalr/).

```{r}
library(portalr)
```

Now that we have loaded the packages we need, we are going to download
some rodent abundance data. This `abundance()` function downloads the
entire dataset, so it might take a few minutes.

```{r include=FALSE}
abund <- abundance(level = "Plot",
                   type = "Rodents")
```

Let's take a look at what the dataset looks like.

```{r}
head(abund)
```

We want to choose data only from a certain data collection period and
only for certain species. To do this, we are going to filter the data to
select trapping period "440" (the 440th time the site was trapped--this
was back in 2015. We are over 500 trapping sessions now!) and get rid of
the rodent removal plots.

We are also going to select only the columns that are relevant to us.

For this assignment, we are going to be calculating values for two
different species of pocket mice: *Chaetodipus penicillatus* and
*Chaetodipus baileyi*. In the Portal data, these species are abbreviated
as `PP` and `PB`, respectively.

When we choose which columns we keep, we will keep the period, plot, and
the PP and PB columns. These columns have the number of each species
caught on each plot that trapping session.

```{r}
# filter period 440, exclude "removal" treatment type, and select "PP" and "PB" column
abund440 <- abund[abund$period == 440 & abund$treatment != "removal", 
                  c("period", "plot", "PP", "PB")]
abund440
```

## Estimating the Mean

The first bit of this assignment is very similar to Module 2 Assignment
1, so if you're feeling stuck with the code, you might want to check
that assignment out.

1.  First, we want to create an object for our sample size. We don't
    know `N` in this scenario, but we do know `n`. Create the object
    `n`, which is the number of sample units that make up our sample.
    This will be the same for both species.

    To do this, use the `nrow()` function, which counts up the number of
    rows in a dataframe. The argument for `nrow()` is the name of the
    dataframe.

    ```{r}
    n <- 
    n
    ```

### *Chaetodipus pencicillatus* (PP) Data

We are going to start by calculating some population parameter estimates
and measures of uncertainty for the PPs.

#### Calculating Estimates of Population Parameters

2.  Create an object called `PP_mean` that contains the sample mean for
    the PPs

```{r}
PP_mean <-
PP_mean
```

3.  Create an object called `PP_var` which contains the sample variance.
    Hint: use the `var()` function.

```{r}
PP_var <- 
PP_var
```

#### Calculating Standard Error for the Mean

Reminder: because we don't know what `N` is, we don't need to use the
finite population correction factor when calculating our standard error
and whatnot.

4.  Calculate the variance of the estimate of the mean for the PPs.

```{r}
PP_var_ybar <- 
PP_var_ybar
```

5.  Calculate the SE of the estimate of the mean.

```{r}
PP_SE <- 
PP_SE
```

#### Calculating Confidence Intervals for the Mean

Hooray, we've now calculated the sample mean, sample variance, and the
standard error! Let's calculate a 95% confidence interval for our sample
mean.

6.  Create an object called `alpha95` to represent our alpha value for a
    95% confidence interval. Because this is a constant, we can assign a
    number to it (take a look back at lecture notes).

```{r}
alpha95 <- 
alpha95
```

7.  Create an object called `df` for degrees of freedom. Degrees of
    freedom in this case are equal to the number of sample units in our
    sample (`n`) minus 1. Use the object `n` in your calculation.

```{r}
df <- 
df
```

We are going to use a function called `qt()` to find the t-value that we
need. Assuming you've created the objects `alpha95` and `df` above,
running this code chunk will give us our t-value, or critical value, for
95% CI.

```{r}
t95 <- qt(1-(alpha95)/2, df)
t95
```

8.  Using `t95` and the standard error that we calculated earlier,
    create an object called `half_width` which contains the half-width
    value for the 95% CI.

```{r}
halfwidth_95 <- 
halfwidth_95
```

9.  Calculate the upper and lower limits of the 95% confidence
    intervals. Use objects you've already created to calculate it.

```{r}
upperCI_95 <- 
lowerCI_95 <- 
  
upperCI_95
lowerCI_95
```

10. What happens when we create 90% CI for the data? Let's calculate
    90% CI. (3 points total)
    
a.  Input the correct value for `alpha90` and then run the code chunk to
    get the new t-value, `t90`.

```{r}
alpha90 <- 
alpha90

t90 <- qt(1-(alpha90)/2, df)
t90
```

b.  Now, using `t90`, calculate the 90% CI. and compare to the 95% CI.
    Which one is wider? Why?

```{r}
upperCI_90 <- 
lowerCI_90 <- 
  
upperCI_90
lowerCI_90
```

c.  Compare the 90% CI to the 95% CI. Which one is wider? Why would
    expect that to be the case?

*Answer:*

### *Chaetodipus baileyi* (PB) Data

Let's do the same thing as above for our *Chaetodipus baileyi* (PB)
data.

#### Calculating Estimates of Population Parameters

11. Create an object called `PB_mean` that contains the sample mean.

```{r}
PB_mean <- 
PB_mean
```

12. Create an object called `PB_var` which contains the sample variance.
    Hint: use the `var()` function.

```{r}
PB_var <- 
PB_var
```

#### Calculating Standard Error for the Mean

13. Start by calculating the variance of the estimate of the mean

```{r}
PB_var_ybar <-
PB_var_ybar
```

14. Calculate the SE of the estimate of the mean.

```{r}
PB_SE <- 
PB_SE
```

#### Calculating Confidence Intervals

15. Create 95% CI for the PB data. You should already have the t-value
    in your environment, so you don't need to go through calculating
    that again.

```{r}

upperCI_95_PB <- 
lowerCI_95_PB <- 
  
upperCI_95_PB
lowerCI_95_PB
```

16. Compare the ranges of the 95% CI for the PP and the PB data. Which
    estimate of the mean seems to be more accurate, based on size of the
    confidence intervals? Why do you think that is (AKA what value is driving that difference)? (2 points)

*Answer:*
