---
title: "Module 5 Lab 2: Spatial Patterns"
author: "Ellen Bledsoe"
date: "2022-11-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Spatial Patterns

## Plot-based Estimates

Larvae of carpenter bees counted in 180 inflorescences of a soap-tree yucca

Counts reported as frequencies (114 zeroes, 25 ones, 15 twos...)

```{r}
x  <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10)     # no. larvae per inflorescence
Fx <- c(114, 25, 15, 10, 6, 5, 2, 1, 1, 0, 1) # no. inflorescences with x larvae
```

Frequencies are convenient to report but a pain to work with, so convert to 180 observations

```{r}
larvae <- rep(x, Fx)
larvae
```

Let's look at the data

```{r}
table(larvae)
```

We cab create a histogram two different ways:

```{r}
hist(larvae)
barplot(table(larvae), xlab = "No. larver per flower (x)", ylab = "Frequency (Fx)")
```

Calculate the Index of Dispersion

```{r}
ID <- var(larvae)/mean(larvae)
ID    # ID > 1, so tends towards clumped (if < 1, then tends towards uniform)
```

Calculate test statistic and df to evaluate whether there is sufficient evidence to reject the null as false and claim the dispersion pattern is clumped rather than random

```{r}
df <- (length(larvae) - 1) # length() provides no. objects in a vector, so sample size (n)
  
chi.sq <- ID * df
chi.sq
```

### VERY IMPORTANT NOTE!

Interpreting the p-values for both the chi-square and normal distributions differ from what is on the PowerPoint slides, so be alert! This is mostly because we are indicating whether we want the lower tail or the upper tail of the distribution.

I don't need you to understand why it is different---what you should know is that in EITHER case, if the p value is LESS than 0.10, then we can say it is significant and accept the alternative hypothesis (either clumped or uniform)

### Calculating the p-value

If n \< 100, use chi-square distribution

```{r}
# P-value for clumped (when ID > 1 and we want upper tail)
pchisq(chi.sq, df, lower.tail = FALSE)

# P-value for uniform (when ID < 1 and we want lower tail)
pchisq(chi.sq, df, lower.tail = TRUE)
```

If n \>= 100, compute d and use normal distribution

```{r}
# adjust the test statistic 
d <- sqrt(2 * chi.sq) - sqrt((2 * df) - 1)
   
# P-value for clumped (when ID > 1 and we want upper tail)
pnorm(d, lower.tail = FALSE)
    
# P-value for uniform (when ID < 1 and we want lower tail)
pnorm(d, lower.tail = TRUE)
```
