---
title: 'Module 5 Assignment 1: Distance Sampling'
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Distance Sampling

## Assignment Details

### Purpose

The goal of this assignment is to understand, apply, and interpret detection probability in estimating densities through distance sampling.

### Task

Write R code to successfully answer each question below or write text to successfully answer the question.

### Criteria for Success

-   Code is within the provided code chunks
-   Code is commented with brief descriptions of what the code does
-   Code chunks run without errors
-   Code produces the correct result
-   Text answers correctly address the question asked

### Due Date

Dec 6 at midnight MST

------------------------------------------------------------------------

## Assignment Questions

All questions are worth 1 point unless otherwise specified.

### Resources

In addition to the slides and practice code, the following might be helpful:

-   [Information on unmarked](http://cran.r-project.org/web/packages/unmarked/index.html)
-   [User's manual](https://cran.r-project.org/web/packages/unmarked/unmarked.pdf)
-   [Using unmarked with distance data](https://cran.r-project.org/web/packages/unmarked/vignettes/distsamp.pdf)

## Set-Up

Some colleagues went out and did some distance sampling! They want to know if "young" saguaros (\< 2 meters tall) or ocotillos are found in higher densities in Saguaro National Park. They need us to get the density estimates for them to answer their question.

First things first, let's load our packages.

```{r}
library(tidyverse)
library(unmarked)
```

## Saguaros (\< 2m)

We will use the saguaro data for our first analyses.

### Data Preparation

1.  Read in the saguaro data; call it `saguaro`.

```{r}
saguaro
```

2.  Make a data frame with transect length (i.e., survey effort)

```{r}
transect_length
```

3.  Plot the distribution of detection distances. Does is look how you expected?

```{r}

```

4.  Create an object called `trunc` that sets the truncation distance to 20m to eliminate extreme observations

```{r}
trunc 
```

5.  Create a vector called `distance_bins` to set "cut points" for distance bins; for the saguaros data, let's do every 4 meters. Be sure to adjust the arguments accordingly!

```{r}
distance_bins
```

6.  Use unmarked function `formatDistData` to count the number of detections in each distance bin for each transect. Save this as an object called `sag_data`. Ignore the *warning* message (but if you get an *error* message, that is different!).

```{r}
sag_data
```

7.  Assemble data into the format required by `unmarked`. Call this unmarked frame `UMF_sag`.

```{r}
UMF_sag
```

8.  Check the distribution of detection distances to be used for analysis to make sure truncation has occurred and the distance bins are what you expect (4 meters)

```{r}

```

### Models

9.  Use the `distsamp` function to create 4 models: half-normal (HN), hazard rate (HR), uniform (Unif), and negative exponential (Exp). (2 points)

```{r}
HN
HR
Unif 
Exp  
```

10. Assemble models into a list to compare their AIC values. Identify the top model from this list (HN or HR or Unif or Exp), which you'll use below.

```{r}
models 
```

11. Which model should we use? Why? (2 points)

    *Answer:*

### Parameter Estimates

12. Use the `backTransform` function to calculate the density estimate.

```{r}

```

13. Calculate the confidence intervals for the density estimate.

```{r}

```

14. Write 2-3 sentences addressing the following questions. (3 points)

-   What does the value in question 12 represent?
-   If we had calculated a density estimate using just our data and not our detection function, would we have under- or overestimated the density of saguaros? Why?

*Answer:*

## Ocotillos

Now, let's do the same thing for the ocotillo data that they collected.

### Data Preparation

15. Read in the ocotillo data and call it `ocotillo`. Remember to use `read.csv` this time (instead of `read_csv`)!

```{r}
ocotillo
```

16. Make your data frame with the transect lengths. Call this `transect_oco`.

```{r}
transect_oco 
```

17. Plot the distribution of detection distances to get an idea of what they look like.

```{r}

```

18. This time around, set truncation distance to 25m; save this as an object called `trunc_oco`. Then create your distance vector (`dinstance_bins_oco`) and set the cut points at every 5 meters.

```{r}
trunc_oco 
distance_bins_oco 
```

19. Use the function `formatDistData` to count the number of detection in each distance bin for each transect. Ignore the warning message. Remember to use the correct data and distance bins.

```{r}
oco_data 
```

20. Assemble data into the format required by unmarked, called an "unmarked frame." Remember to refer to the correct data frames and values! Save this as `UMF_oco`

```{r}
UMF_oco 
```

### Models

21. Fit the four models to the ocotillo data. Again, be sure you're using the correct data frame. (2 points)

```{r}
HN_oco
HR_oco 
Unif_oco 
Exp_oco
```

22. Assemble models into a list to compare their AIC values.

```{r}
models_oco 
```

23. Using the best model, calculate the density estimate and confidence intervals for ocotillos. (2 points)

```{r}

```

## Compare Densities

We've done a lot of calculations just to get to this point!

24. Which has the higher density per hectare: saguaros or ocotillos? How do you know?

*Answer:*
