---
title: "Module 2 Assignment 1"
author: ""
date: '2022-10-11'
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Simple Random Sampling -- Rocks!

## Assignment Details

### Purpose

The goal of this assignment is to use R to calculate and understand
accuracy of estimates (standard error specifically).

### Task

Write R code to successfully answer each question below or write text to
successfully answer the question.

### Criteria for Success

-   Code is within the provided code chunks
-   Code chunks run without errors
-   Code produces the correct result
    -   Code attempts will get half points
    -   Code that produces the correct answer will receive full points
-   Text answers correctly address the question asked

### Due Date

October 10 by 1 pm MST

------------------------------------------------------------------------

## Assignment Questions

### Short Answer

Think back to how we sampled the rock data. Answer the following
question in terms of our data (aka, "big square," "little squares,"
"dark rocks," etc.)

1.  Define the *sample* standard deviation of the mean in terms of our
    data (not the equation but what we use to calculate standard
    deviation, what it tells us, and what it is an estimate of).

    *Answer:*

2.  Define the standard error of the estimate of the mean (again, in
    terms of our data and what it tells us).

    *Answer:*

3.  How do sample standard deviation and standard error differ? (If
    you've explained that above, reiterate it here.)

    *Answer:*

4.  What would happen to the sample standard deviation if we sampled 20
    sampling units (little squares) instead of only 8? Would it
    increase, decrease, or stay about the same? How about the standard
    error? Explain your rationale for both.

    *Answer:*

### Calculations

Each question is worth 1 point unless otherwise noted.

#### Set-up:

(a) Get your (or your group's) simple random sampling rock data.
(b) Make sure it has been saved as a .csv file. If it hasn't, you will
    need to do so using "Save As".
(c) In RStudio Cloud, using the "Upload" button in the lower right-hand
    panel, upload your .csv file.
    (d) In the Files tab in the bottom right-hand corner, check the box
        next to the .csv file that you just uploaded, then click the
        "Rename" button (near "Upload").
(d) Rename your .csv file to "rock_data.csv"

Load the tidyverse. For this assignment, we will *only* be using this to
read in our .csv file.

```{r}
library(tidyverse)
```

5.  Use the `read_csv()` function from the tidyverse to read in your
    rocks file. Save this data in an object called `rocks`.

    Below, replace the file name with the file name of your rock data.

```{r}
rocks <- read_csv("your_file_name_here.csv") %>% drop_na()
```

6.  Create the object `N`, which is the number of units in your
    *sampling frame*.

```{r}
N <- 
```

7.  Create the object `n`, which is your *sample size,* the number of
    sample units that make up your sample. To do this, use the `nrow()`
    function

    (Hint: if you don't remember how to use the `nrow()` function, run
    `?nrow`)

```{r}
n <- 
```

### Estimating the Mean

One way we can calculate values is to call the column that we want and
then save that value to an object.

For example, let's say we wanted to calculate the median of our dark
stone abundance data. We could run the following line of code:
`rocks_median <- median(rocks$abund_dark_stones)`

8.  Create an object called `rocks_mean` that contains the sample mean
    (AKA the mean number of dark stones per sample unit). Follow the
    code example above.

```{r}
rocks_mean <- 
```

9.  Create objects called `rocks_var` and `rocks_sd` which contain the
    sample variance and sample standard deviation, respectively. Hint:
    use `var()` and `sd()` functions.

```{r}
rocks_var <- 
rocks_sd <- 
```

### Calculating Standard Error for the Mean

R doesn't have a function for calculating the standard error (SE), so
let's calculate it ourselves using the equation from class.

10. First, do you remember that finite population correction factor?
    Let's create an object called `pop_correction` using that formula.

    (Hint: use the objects we've already created (`N`, `n`; don't enter
    any numbers directly)

```{r}
pop_correction <- 
```

11. Why are we using the finite population correction factor in our
    formulas?

    *Answer:*

12. Using the objects you've already created and the formula (check the
    lecture slides from this module!), create an object called
    `rocks_var_ybar` which contains the variance of the *estimate* of
    the mean.

    Remember, this is different from the sample variance that you've
    already calculated, but that sample variance is included...

```{r}
rocks_var_ybar <- 
```

13. Calculate the SE of the estimate of the mean by taking the square
    root of the `rocks_var_ybar` value. Call this object `rocks_SE`.

```{r}
rocks_SE <- 
```

### Estimating the Population (and Accuracy of the Estimate)

We've calculated standard error (SE) for the estimate of the mean above.
Now, let's calculate them for the estimate of the population abundance
(tau hat)

14. First, let's calculate the total number of rocks in the population.
    Call this `rocks_total`. Use objects you've already created.

```{r}
rocks_total <- 
```

15. Now calculate the variance of the estimate of the population total,
    `rocks_var_total`.

```{r}
rocks_var_total <- 
```

16. Calculate the SE of the estimate of the population total,
    `rocks_SE_total`.

```{r}
rocks_SE_total <- 
```

### Bringing It All Together

17. Write a short paragraph (3-5 sentences) putting the values that
    we've just calculated into context for our dark rocks sampling
    adventure. Be sure to answer the following questions. (3 points)

-   According to your calculations, approximately how many dark rocks
    were there in your big square? How confident are you in that
    estimate?
-   What is your estimate of density of dark rocks (average \# dark
    rocks per little square)? How much variation do you estimate is
    there in the number of dark rocks per little square? How confident
    are you in your estimate of the density?

*Answer:*
