---
title: "Module 2, Assigment 1"
author: "Ellen Bledsoe" # <- your name here
date: '2022-09-19'      # <- current date here
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Purpose

The goal of this assignment is to explore computational reproducibility and apply the coding skills we've learned and practiced in class and lab.

### Task

Write R code to successfully answer each question below and/or write text to successfully answer the question.

### Criteria for Success

-   Code is within the provided code chunks
-   Code chunks run without errors
-   Code produces the correct result
    -   Code attempts will get half points
    -   Code that produces the correct answer will receive full points
-   Text answers correctly address the question asked

### Due Date

Feb 20 at midnight MST

# Assignment Questions

This assignment is worth 20 points total

## Definitions (1 point each)

*In your own words*, define/describe the following terms. These don't need to be technical descriptions but rather how you are thinking about them.

1.  *Reproducibility*:

2.  *Open science*:

3.  *R*:

4.  *RStudio*:

## Vectors (1 point each)

5.  Run the line below to create the object heights. Use the `mean()` function to calculate the mean value of height.

```{r}
heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63, NA, 72, 65, 64, 70, 63, 65)
```

```{r}

```

6.  This should yield an odd result caused by the `NA` values. To resolve this, use the help function to remind yourself about the argument `na.rm = TRUE` that applies to many R functions

```{r}
help(mean) # same as ?mean
```

Issue a revised command to calculate the mean value in `heights` and send the result to the console.

```{r}

```

7.  Write a line of code that selects the 6-10th height values.

```{r}
heights[6:10]
```

8.  Create a character vector called `rhymes` that contains the following values: cat, rat, bat, hat

```{r}

```

## Data Frames

**IMPORTANT!** Run the following code chunk to read in the cactus pad dataset. *You don't need to understand what is happening, though I've added some comments in case you are interested!*

We will be using that data for the remainder of the assignment.

Once you run the code chunk, the data frame will be saved in your environment as the object `pads`.

```{r}
# read in the cactus pads data
pads <- read.csv("../data_raw/CactusPadsData.csv") 

# read in the cactus data and choose only the group_id, species, and size columns
cactus <- read.csv("../data_raw/CactusData.csv") |> 
  dplyr::select(group_id, species, size)

# join the two datasets together using the group_id column
# this matches up the cactus species and size for each cactus pad (row) with the same group_id
pads <- dplyr::right_join(pads, cactus, by = "group_id")

# remove the cactus data frame because we don't need it anymore (and it might be confusing)
rm(cactus)
```

9.  How many rows does this data frame have? How many columns? You can either use code to figure this out or look at the object in the environment.

    *Rows*: 
    
    *Columns*:

```{r}
# optional space for code to answer the question above
```

10. Look at the first 6 rows of data. You can either do this by using a function or by using index sub-setting.

```{r}

```

11. Calculate the mean of width of all the cactus pads that were measured.

```{r}

```

12. Create a histogram with the length of all the cactus pads that were measured.

```{r}

```

20. What does the following line of code do?

```{r}
surveys[surveys$species_id == "DM", ]
```

*Answer:*

## Calculating the Coefficient of Variance

The remainder of the questions in this assignment lead you through the steps to calculate the coefficient of variance for two different species at the field site: the largest (NL, the woodrat *Neotoma albigula*) and one of the smallest (BA, the Northern pygmy mouse, *Baiomys taylori*).

21. Create 2 different data frames: one called `NL` with rows that have "NL" in the species_id column and one called `BA` with only rows which have "BA" in the species_id column. (Hint: take a look at question 20 for inspiration on how to do that)

```{r}
NL <- 
BA <- 
```

22. Calculate the mean weight for each species.

```{r}
mean_NL <- 
mean_BA <- 
```

23. In a similar fashion, calculate the standard deviation (`sd()`) for the weight of both species.

```{r}
sd_NL <- 
sd_BA <- 
```

24. Using the objects you just created in questions 22 and 23, calculate the coefficient of variance (CV) for each species.

```{r}
cv_NL <- 
cv_BA <- 
```

25. In 1-2 sentences, interpret the results of your calculations.

*Answer:*

## Assignment Submission Guidelines

When you are finished, I want you to export this .Rmd file and upload it into the assignment dropbox on D2L.

To do you, follow these steps:

-   Make sure you have saved all of your code!
-   In the "Files" pane, check the empty box near this .Rmd file
-   Still in the "Files" pane, click the small downward arrow next to the gear ("More")
-   Select "Export"
-   Rename you file with the following pattern: "LastName_Module2_Assignment1.Rmd"
-   Once you've exported, the .Rmd file will be in your "Downloads" folder on your computer
-   Submit this file to D2L
