---
title: "Module 2 Assigment 1"
author: "Ellen Bledsoe"
date: '2022-09-19'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Definitions (1 point each)

_In your own words_, define/describe the following terms. These don't need to be
technical descriptions but rather how you are thinking about them.

Grading note: I've copied these definitions from my slides. If their
definitions are the same, they have copied directly from the slides and
not used their own words, so take 0.5 points off.

1.  Reproducibility: ability to repeat the original study using the same
    data, materials, and methods

2.  Open science: Scientific research conducted and communicated in an
    honest, accessible, and transparent way, such that--at a minimum--a
    study can be reproduced and/or replicated.

3.  R: R refers to both the programming language and the software that
    interprets scripts written in the language.

4.  RStudio: RStudio is an integrated development environment (IDE) that
    helps us interact with R more easily.

## Short-Answer Question (2 points)

5.  In 2-3 sentences, explain why data management is important for
    reproducibility.

Key points they should touch on (at least some): 

* best practices help with organization
* also important for effective communication (to future you, your colleagues, and others who might want to reproduce your
work) 
* makes transparency higher

## Vectors (1 point each)

6.  Multiply 8 x 5.
```{r}
8 * 5
```


7.  Create the object x and have it contain the result of 40 / 2.77 
```{r}
x <- 40 / 2.77
x
```

8.  Create the object y and have it contain the result of the square
    root of 9 (HINT: use the `sqrt` function).
```{r}
y <- sqrt(9)
y
```

9.  Create the object num_vec that contains the values 1, 3, 5, 7
```{r}
num_vec <- c(1, 3, 5, 7)
num_vec
```


10. Run the line below to create the object heights. Then add the value
    80 to the end of the vector. Keep the same name for the vector
    (heights). Note: NA is the value used by R to identify missing data.

heights \<- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63,
NA, 72, 65, 64, 70, 63, 65)

heights \<- c(heights, 80) if they do not rename the vector (AKA only
have c(heights, 80), half a point off)

11. Use the mean() function to calculate the mean value of height and
    send the result to the console mean(heights)

12. This should yield an odd result caused by the NAs. To resolve this,
    use the help function to learn about the argument na.rm = TRUE that
    applies to many R functions

```{r}
help(mean) # same as ?mean
```

Issue a revised command to calculate the mean value in num_vec and send
the result to the console. mean(heights, na.rm = TRUE) \# na.rm = T is
also valid

13. Write and execute a line of code that selects the 6-10th height
    values 
```{r}
heights[6:10]
```

14. Execute the line below. What does this line of code do?

```{r}
heights67 <- heights[heights > 67]
```

Answer: selects values from the heights vector that are greater than 67 and assigns this vector to a new object

15. Execute the two lines below. What do these lines of code tell us?

```{r}
length(heights) 
length(heights67)
```
Answer: The heights vector has 22 values but only 9 of them are greater than 67

16. Create a character vector called `rhymes` that contains the following values: cat, rat, bat, hat 
```{r}
rhymes <- c("cat", "rat", "bat", "hat")
```

17. Run the lines below. What does each line do?

char_vec == 'C' char_vec != 'C'
```{r}
rhymes == "hat"
rhymes != "hat"
```

Answer: First line creates a logical vector with "TRUE" equaling "hat." The second line also creates a logical vector for
all values _except_ "hat"

## Data Frames

Run the following line of code. Like the Palmer Penguins dataset that we used in class, `ratdat` is also a data package. It has data from my PhD field site, the Portal Project. You can learn more about the Portal Project [here](https://portal.weecology.org/).

We are going to be using the `surveys` dataset from the `ratdat` package.
```{r}
library(ratdat)
surveys
```


18. Look at the first 6 rows of data. You can either do this by using a
    function or by using index subsetting (with the square brackets: [])
```{r}
head(surveys) 
surveys[1:6,] # either one is fine
```


19. How many rows does this data frame have? How many columns? 
Rows: 34786 
Columns: 13

20. Use a function to print the column names of surveys.
```{r}
names(surveys)
colnames(surveys) # either one is fine
```

21. Create a histogram with the hindfoot lengths column
```{r}
hist(surveys$hindfoot_length)
```


22. What does the following line of code do? 
```{r}
surveys[surveys$species_id == "DM", ]
```

Answer: selects rows from surveys in which the value in the species_id column is "DM"

23. Explain what the following lines of code do (2 points). 
```{r}
weights_noNA <- surveys[!is.na(surveys$weight), ] 
weights_over200g <- weights_noNA[weights_noNA$weight > 200, ]
```

Answer 
First line: Selects rows from surveys in which the weight column
is not NA values (so only numeric values) and creates an object called
weights_noNA 
Second line: Selects rows from the weights_noNA object in
which the values in the weight column are greater than 200, saving them
in an object called weights_over200g



When you are finished, I want you to export this .Rmd file and upload it
into the assignment dropbox on D2L.

To export the file, do the following: save this script as
"LastName_Module2_Assignment1.Rmd"
