---
title: "Introduction to the `tidyverse`"
author: "Ellen Bledsoe"
date: "2022-09-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What is the `tidyverse`? 

The `tidyverse` is an umbrella package, or a collection of packages! All packages included in the tidyverse share the same underlying structure, syntax (grammar), and data structures. They are designed to work seemlessly together.

Everything we have worked with until now has been part of base R, or functions and language that come already installed with R. The `tidyverse` works a bit differently but is actually more intuitive!

To learn more about the packages in the tidyverse, check out these links:

* [List of packages and what they do](https://www.tidyverse.org/packages/)
* [RStudio Cheatsheets](https://www.rstudio.com/resources/cheatsheets/)
  - specifically the `dplyr` and `tidyr` cheatsheets, but there are many more!
* [Data Carpentry lesson](https://datacarpentry.org/R-ecology-lesson/03-dplyr.html)
* [Effectively using the tidyverse](https://r4ds.had.co.nz/index.html)

## Set Up

I have already set up these RStudio Cloud projects so that the `tidyverse` is installed. Each time we want to use the `tidyverse`, however, we need to "load" it into our workspace. We do this with the `library()` function.

We are also going use the `penguins` dataset from the `palmerpenguin` package. We will need to load that package, as well.

```{r}
library(tidyverse)
library(palmerpenguins)
```

When you load the `tidyverse` library, you'll see some "conflicts." Don't panic! That's normal. Those conflicts are telling us that certain functions in `dplyr` are now overriding some functions in base R with the same name.

## Penguins

# Let's remind ourselves what the `penguins` dataset looks like.

```{r}
head(surveys)
colnames(surveys)
str(surveys)
view(surveys)

# glimpse() is from the dplyr package
glimpse(surveys)
```

## SELECT and FILTER #

# use select() to subset columns
# use filter() to subset rows based on conditions

# the first argument is the dataset; all following arguments are the column names
select(surveys, plot_id, species_id, weight)

# to select columns *except* certain ones, use the minus sign 
select(surveys, -record_id, -species_id)

# to choose rows based on specific criteria, use filter()
# first argument is the data; all subsequent arguments are selection criteria
filter(surveys, year == 1995)
filter(surveys, year >= 1995)
filter(surveys, species_id == 'DM')
filter(surveys, year >= 1995, species_id == 'DM')

#---------------------------------------------------------------------------####
# COMBINING FUNCTIONS WITH PIPES #

# the pipe operator is %>% 
# it's originally from the `magrittr` package but tidyverse packages load it, too
# it helps us string together multiple functions without a bunch of intermediate
# steps in between that can become confusing!

# Shortcut: Ctrl + Shift + M (Windows) or Cmd + Shift + M (Mac)

# Say we want to filter the data and only select certain columns. We can do that
# by creating a temporary dataframe. That gets cumbersome quickly, though! 
surveys2 <- filter(surveys, weight < 5)
surveys_sml <- select(surveys2, species_id, sex, weight)

rm(surveys_sml)

# Alternatively, we can use the pipe.
surveys_sml <- surveys %>% 
  filter(weight < 5) %>% 
  select(species_id, sex, weight)

# Challenge #
# Using pipes, subset the surveys data to include animals collected before 1995
# and retain only the columns year, sex, and weight.

surveys %>% 
  filter(year < 1995) %>% 
  select(year, sex, weight)

# why didn't this show up in the environment?
# because we didn't assign the output to an object!

#---------------------------------------------------------------------------####
# MUTATE #

# the mutate() function helps us create new columns from existing columns and
# applying a mathematical function or other functions 

surveys %>% 
  mutate(weight_kg = weight / 1000)

# we can add head() or glimpse() or other viewing functions to confirm!
surveys %>% 
  mutate(weight_kg = weight / 1000,
         weight_lb = weight_kg * 2.2) %>% 
  glimpse()

# there are lots of NAs here! let's remove them with the filter function
surveys %>% 
  filter(!is.na(weight)) %>% 
  mutate(weight_kg = weight / 1000) %>% 
  glimpse()


#---------------------------------------------------------------------------####
# GRIOUP BY and SUMMARIZE #

# often, we need to follow the "split-apply-combine" paradigm in data analysis:
#   - split data into groups
#   - apply some type of calculation or analysis to each group
#   - combine the results
# we can use group_by() and summarize() functions from `dplyr` to do this 

# the group_by() function splits the df up into sections by categorical data
# the summarize() function creates a new column with the new calculations
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))

# let's filter out the NAs at the beginning
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight))

# we can run multiple calculations in the summarize function, as we did with mutate
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            sd_weight = sd(weight))

# let's arrange the resulting dataframe from smallest mean weight to greatest
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            sd_weight = sd(weight)) %>% 
  arrange(mean_weight)

# we can add the descending argument to go from greatest to smallest
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            sd_weight = sd(weight)) %>% 
  arrange(desc(mean_weight))

# COUNT #

# Another way to summarize data is through the count() function, which tells us
# how many observations are in each group

surveys %>% 
  count(species_id)

# let's combine with arrange
surveys %>%
  count(sex, species_id) %>%
  arrange(species_id, desc(n))
