---
title: "Module 2, Assignment 2"
author: ""            # <- your name here!
date: '2022-09-27'    # <- you can change the date if you want
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment Details

### Purpose

The goal of this assignment is to use and become comfortable with working with 2-dimensional data using the `tidyverse`.

### Task

Write R code to successfully answer each question below or write text to successfully answer the question.

### Criteria for Success

-   Code is within the provided code chunks
-   Code is commented with brief descriptions of what the code does
-   Code chunks run without errors
-   Code produces the correct result
-   Text answers correctly address the question asked

### Due Date

Sept. 29 at midnight MST

# Assignment Questions

## Definitions (2 point each)

*In your own words*, define/describe the following terms. These don't need to be technical descriptions but rather how you are thinking about them.

1.  *data frame/tibble*:

2.  *R package*:

3.  *tidyverse*:

4.  *the pipe (`%>%`)*:

## Short-Answer Question (2 points)

5.  What are 2 reasons why the pipe is useful? Answers can be in sentence or bullet-point form.

*Answer*:

## Using the `tidyverse`

For the rest of the assignment, you will be writing or interpreting code from the `tidyverse` and using data from the `surveys` data frame in the `ratdat` package.

6.  Load the `tidyverse` package and the `ratdat` package into RStudio.

```{r}

```

### *Optional*

The `surveys` data frame is part of the `ratdat` package. Since we have loaded in the `ratdat` package, we can access the `surveys` data frame even though we can't see it in our environment. If you want to see the `surveys` data frame in your environment, we need to give it a name by running the following line of code. You do not *need* to do this, but you're welcome to do so if you want.

```{r}
surveys <- surveys
```

7.  There are many different functions we can use to look at our data set. Choose your favorite to get an idea of all the data in `surveys`.

```{r}

```

8.  Which of the columns in `surveys` have data which are character data types? List the columns below.

    *Answer:*

9.  *In your own words*, explain what the `select()` function (from tidyverse) does.

    *Answer:*

10. Use the `select()` function to create a data frame that has only columns with the following data: year, plot number, species, weight. (Hint: make sure you are typing the correct column names!)

```{r}

```

11. *In your own words*, explain what the `filter()` function does.

    *Answer:*

12. Create a data frame from `surveys` that only includes observations from the year 2000 and later.

```{r}

```

13. Now create a data frame that only includes observations from before the year 1990 and the species_id "BA".

```{r}

```

14. Run the lines of code below.

```{r}
krats <- surveys %>% 
  filter(species_id == "DM" | species_id == "DO" | species_id == "DS") %>% 
  select(year, plot_id, species_id, hindfoot_length)
```

Describe what each of the three lines are doing (3 points). (Hint: the `|` means "or")

*Line 1*:\
*Line 2*:\
*Line 3*:

15. *In your own words*, explain what the `mutate()` function does.

    *Answer:*

16. The hindfoot_length column in currently in mm. Create a new column named `hindfoot_length_cm` where the hindfoot measurement is now in centimeters. (Hint: to convert from mm to cm, divide by 10)

```{r}

```

17. *In your own words*, explain what the `group_by()` function does.

    *Answer:*

18. For this question, we are going to use `krats`, the data frame you created in question 14.

    Use the `group_by()` and `summarize()` functions together to calculate the mean hindfoot_length (mm) for each species in `krats`. (2 points)

    Remember to use the `%>%`, and remember to exclude `NA`s from the mean calculation.

```{r}

```

19. We are back to using the `surveys` data frame for this question.

    Create a new data frame called `PE_weights`. It should have the minimum (`min()`) and maximum (`max()`) weight of the "PE" species for each year. (4 points)

This question has a lot of steps, so think through each one individually before trying to write the code.

-   which function will you need to use to get only "PE" individuals?
-   which function will you need to get values for *each* year?
-   remember to use the `na.rm` argument to return numbers

```{r}

```

## Assignment Submission Guidelines

When you are finished, I want you to export this .Rmd file and upload it into the assignment dropbox on D2L.

To do you, follow these steps:

-   Make sure you have saved all of your code!
-   In the "Files" pane, check the empty box near this .Rmd file
-   Still in the "Files" pane, click the small downward arrow next to the gear ("More")
-   Select "Export"
-   Rename you file with the following pattern: "LastName_Module2_Assignment2.Rmd"
-   Once you've exported, the .Rmd file will be in your "Downloads" folder on your computer
-   Submit this file to D2L
