---
title: "Module 3 Assignment 3"
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Distance Sampling

## Assignment Details

### Purpose

The goal of this assignment is to understand, apply, and interpret detection probability in estimating densities through distance sampling.

### Task

Write R code to successfully answer each question below or write text to successfully answer the question.

### Criteria for Success

-   Code is within the provided code chunks
-   Code chunks run without errors
-   Code produces the correct result
    -   Code attempts will get half points
    -   Code that produces the correct answer will receive full points
-   Text answers correctly address the question asked

### Due Date

April 24 at 2pm MST

------------------------------------------------------------------------

## Assignment Questions

All questions are worth 1 point unless otherwise specified.

### Resources

In addition to the slides and practice code, the following might be helpful:

-   [Information on unmarked](http://cran.r-project.org/web/packages/unmarked/index.html)
-   [User's manual](https://cran.r-project.org/web/packages/unmarked/unmarked.pdf)
-   [Using unmarked with distance data](https://cran.r-project.org/web/packages/unmarked/vignettes/distsamp.pdf)

## Set-Up

First things first, let's load our packages.

```{r message = FALSE}
library(tidyverse)
library(unmarked)
```

## Saguaros (\< 2m)

We will use the saguaro data for our first analyses.

### Data Preparation

1.  Read in the saguaro data; call it `saguaro`.

```{r}
beads <- read.csv("../data_raw/Distance_Data.csv")
```

2.  Make a data frame with transect length (i.e., survey effort)

```{r}
transect_length <- beads %>% 
  group_by(GroupID) %>% 
  summarise(length = mean(TransectLength_ft))
```

Before continuing our analyses, we will want to create one dataframe for the clear beads and another dataframe for the green beads. You can do this either through base R or the `tidyverse`...both will use conditional statements!
```{r}
clear <- beads %>% 
  filter(BeadColor == "Clear")

green <- beads %>% 
  filter(BeadColor == "Green")
```

Take a look at the number of rows in two data frames that you just created. What do you notice? Why do you think that is?

### Clear Beads
3.  Plot the distribution of detection distances. Does is look how you expected?

```{r}
hist(clear$Distance_in) 
```

4.  Create an object called `trunc` that sets the truncation distance to 20m to eliminate extreme observations

```{r}
trunc <- max(clear$Distance_in)
```

5.  Create a vector called `distance_bins` to set "cut points" for distance bins; for the saguaros data, let's do every 12 inches (every foot). Be sure to adjust the arguments accordingly!

```{r}
distance_bins <- seq(0, trunc, by = 12) 
```

6.  Use unmarked function `formatDistData` to count the number of detections in each distance bin for each transect. Save this as an object called `sag_data`. Ignore the *warning* message (but if you get an *error* message, that is different!).

```{r}
clear_data <- formatDistData(clear, 
                             distCol = "Distance_in", 
                             transectNameCol = "GroupID", 
                             dist.breaks = distance_bins)
```

7.  Assemble data into the format required by `unmarked`. Call this unmarked frame `UMF_sag`.

```{r}
UMF_clear <- unmarkedFrameDS(y = as.matrix(clear_data), 
                           survey = "line",
                           tlength = transect_length$length, 
                           dist.breaks = distance_bins,
                           unitsIn = "in")
```

8.  Check the distribution of detection distances to be used for analysis to make sure truncation has occurred and the distance bins are what you expect (4 meters)

```{r}
hist(UMF_clear)
```

### Models

9.  Use the `distsamp` function to create 4 models: half-normal (HN), hazard rate (HR), uniform (Unif), and negative exponential (Exp). (2 points)

```{r}
HN   <- distsamp(~1 ~1, UMF_clear, keyfun = "halfnorm", output = "density", unitsOut = "ha")
HR   <- distsamp(~1 ~1, UMF_clear, keyfun = "hazard", output = "density", unitsOut = "ha")
Unif <- distsamp(~1 ~1, UMF_clear, keyfun = "uniform", output = "density", unitsOut = "ha")
Exp  <- distsamp(~1 ~1, UMF_clear, keyfun = "exp", output = "density", unitsOut = "ha")
```

10. Assemble models into a list to compare their AIC values. Identify the top model from this list (HN or HR or Unif or Exp), which you'll use below.

```{r}
models <- fitList('Half Normal' = HN, 
                  'Hazard Rate' = HR,
                  'Uniform'     = Unif,
                  'Exponential' = Exp)
modSel(models)
```

11. Which model should we use? Why? (2 points)

    *Answer:* Half-normal, lowest AIC value

### Parameter Estimates

12. Use the `backTransform` function to calculate the density estimate.

```{r}
backTransform(Exp, type="state") # Density estimate (no./ha) 
```

13. Calculate the confidence intervals for the density estimate.

```{r}
exp(confint(Exp, type="state"))  # CI for density
```

14. Write 2-3 sentences addressing the following questions. (3 points)

-   What does the value in question 12 represent?
-   If we had calculated a density estimate using just our data and not our detection function, would we have under- or overestimated the density of saguaros? Why?

*Answer:* number of saguaros \<2m per hectare; under because we could not account for the ones we missed during sampling

## Ocotillos

Now, let's do the same thing for the ocotillo data that they collected.

### Data Preparation

15. Read in the ocotillo data and call it `ocotillo`. Remember to use `read.csv` this time (instead of `read_csv`)!

```{r}
ocotillo <- read.csv("../data_raw/ocotillo.csv")
```

16. Make your data frame with the transect lengths. Call this `transect_oco`.

```{r}
transect_length_oco <- ocotillo %>% 
  group_by(group) %>% 
  summarise(length = mean(length))
```

17. Plot the distribution of detection distances to get an idea of what they look like.

```{r}
hist(ocotillo$distance) 
```

18. This time around, set truncation distance to 25m; save this as an object called `trunc_oco`. Then create your distance vector (`dinstance_bins_oco`) and set the cut points at every 5 meters.

```{r}
trunc_oco <- 25
distance_bins_oco <- seq(0, trunc_oco, by = 5) 
```

19. Use the function `formatDistData` to count the number of detection in each distance bin for each transect. Ignore the warning message. Remember to use the correct data and distance bins.

```{r}
oco_data <- formatDistData(ocotillo, 
                           distCol = "distance", 
                           transectNameCol = "group", 
                           dist.breaks = distance_bins_oco)
```

20. Assemble data into the format required by unmarked, called an "unmarked frame." Remember to refer to the correct data frames and values! Save this as `UMF_oco`

```{r}
UMF_oco <- unmarkedFrameDS(y = as.matrix(oco_data), 
                           survey = "line",
                           tlength = transect_length_oco$length, 
                           dist.breaks = distance_bins_oco,
                           unitsIn = "m")
UMF_oco
```

### Models

21. Fit the four models to the ocotillo data. Again, be sure you're using the correct data frame. (2 points)

```{r}
HN_oco   <- distsamp(~1 ~1, UMF_oco, keyfun = "halfnorm", output = "density", unitsOut = "ha")
HR_oco <- distsamp(~1 ~1, UMF_oco, keyfun = "hazard", output = "density", unitsOut = "ha")
Unif_oco <- distsamp(~1 ~1, UMF_oco, keyfun = "uniform", output = "density", unitsOut = "ha")
Exp_oco  <- distsamp(~1 ~1, UMF_oco, keyfun = "exp", output = "density", unitsOut = "ha")
```

22. Assemble models into a list to compare their AIC values.

```{r}
models_oco <- fitList('Half Normal' = HN_oco, 
                      "Hazard Rate" = HR_oco,
                      'Uniform'     = Unif_oco,
                      'Exponential' = Exp_oco)
modSel(models_oco)
```

23. Using the best model, calculate the density estimate and confidence intervals for ocotillos. (2 points)

```{r}
backTransform(HN_oco, type="state") # Density estimate (no./ha) 
exp(confint(HN_oco, type="state"))  # CI for density
```

## Compare Densities

We've done a lot of calculations just to get to this point!

24. Which has the higher density per hectare: saguaros or ocotillos? How do you know?

*Answer:*
