---
title: "Module 3 Assignment 2"
author: "Ellen Bledsoe"
date: "2022-10-18"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Stratified Random Sampling

## Assignment Details

### Purpose

The goal of this assignment is to understand, apply, and interpret a
stratified random sampling method and calculations.

### Task

Write R code to successfully answer each question below or write text to
successfully answer the question.

### Criteria for Success

-   Code is within the provided code chunks
-   Code is commented with brief descriptions of what the code does
-   Code chunks run without errors
-   Code produces the correct result
-   Text answers correctly address the question asked

### Due Date

October 20 at midnight MST

------------------------------------------------------------------------

## Assignment Questions

We have been asked to go out and sample for grasshoppers. The location
we are sampling has 3 different vegetation types: riparian, shrubland,
and grassland. We have decided to use a stratified random sample design
for our survey.

1.  In about 2 sentences, explain why we have chosen stratified random
    sampling as apposed to simple or systematic? What benefit does it
    provide us? (2 points)

    *Answer:*

### Set-up

2.  Load the `tidyverse` package. We won't be using it *too* much this
    time around, but we need it a couple times.

```{r}
library(tidyverse)
```

### Part 1: Allocation of Sample Units

Let's say we have a sampling frame with 1000 sample units. We can sample
100 of those units. We have our 3 strata which we have determined via
vegetation type (we will called them A, B, and C for simplicity for the
rest of the assignment).

How do we decide how many sampling units to sample from each stratum?

3.  Create an object, `N`, which represents the total number of sample
    units in our *sampling frame*. We also need to create `n`, which is
    the total number of sample units in our *sample*.

```{r}
N <- 
n <- 
```

We have determined that Stratum A has 200 sampling units, Stratum B has
500 sampling units, and Stratum C has 300 sampling units.

4.  Add the correct values to the lines of code below to create objects
    for future use.

```{r}
N_A <-
N_B <-
N_C <-
```

Due to a pilot study done last year, we have an estimate of the amount
of variation (standard dev.) for each stratum.

```{r}
sd_A <- 9
sd_B <- 4
sd_C <- 7
```

We also have an estimate of the time-cost (hours) it takes to survey
each unit.

```{r}
cost_A <- 1.5
cost_B <- 2
cost_C <- 4
```

#### Optimal Allocation

We want to use the optimal allocation equation to calculate how many
sample units each stratum should have. Let's do this incrementally.

5.  Start by calculating the N~i~ \* s~i~ / square root of c~i~ for each
    stratum. Remember to use objects you've created rather than typing
    out specific numbers into the equation.

```{r}
opt_allo_A <-
opt_allo_B <-
opt_allo_C <-
```

6.  Calculate the denominator of the equation by adding all of the
    values above together.

```{r}
opt_allo_total <- 
```

7.  Calculate how many sampling units we should optimally sample in each
    stratum. Use the `round()` function to round to the nearest whole
    number.

```{r}
n_A <- 
n_B <- 
n_C <- 
```

8.  If we did not have the pilot study telling us about the variation of
    the populations in each strata, we could have allocated very
    differently:

    -   stratum A = 24

    -   stratum B = 53

    -   stratum C = 23

    Why is this? What would this allocation have meant for the precision
    of our estimates? (2 points)

    *Answer:*

### Part II: Stratified Random Sampling

#### Data

Hooray! We have gone out and sampled the number of grasshoppers per
sample unit, allocating our sample units optimally (including variation
in population).

Now we want to calculate our population parameters:

-   mean grasshoppers per sampling unit (`y_bar`)
-   total grasshoppers in the population (`tau_hat`)

9.  Let's load our "grasshoppers.csv" data into our workspace using the
    `read_csv()` function. Save the data frame as an object called
    `grasshoppers`.

```{r}

```

10. Explore the dataset (either through code or in the environment
    panel).

    What does each column represent? What does one row represent? (2
    points)

    *Answer:*

```{r}

```

11. Run the following code chunk. What do these numbers represent?

    *Answer:*

    ```{r}
    grasshoppers %>% 
      group_by(stratum) %>% 
      count(stratum)
    ```

    (Hint: do these numbers match your values from question 7? If not,
    you probably want to revisit your calculations up in the previous
    section!)

Based on what I've taught you in R thus far, the easiest way to do our
calculations will be to have three different data frames, one for each
stratum.

```{r}
stratumA <- grasshoppers %>% filter(stratum == "A")
stratumB <- grasshoppers %>% filter(stratum == "B")
stratumC <- grasshoppers %>% filter(stratum == "C")
```

That said, if you're feeling super ambitious and want to *try* to use
the `tidyverse` and a lot of `group_by()`, `mutate()`, and `summarise()`
functions, be my guest! To be clear, though, that is neither required
nor expected.

#### Calculations

NOTE: Open up your script from Module 3, Assignment 1---it's about to
come in handy! I would also recommend opening up the lecture slides from
lecture 14.

12. Before we begin calculating, stop for a moment to reflect and
    predict. Take a minute to skim through the lecture slides and remind
    yourself of the process of how we go about calculating the *overall*
    population mean and total abundance when we use stratified sampling.
    Describe (either in sentences or in bullet points) how you think we
    will proceed.

    Note: This question in graded for completion only, not for accuracy.
    (2 points)

    *Answer:*

#### Step 1: Calculating Sample Statistics (per stratum)

For our first step, we will calculate for each stratum the mean number
of grasshoppers, the total number of grasshoppers, and the sample
variances.

13. Use the code from Module 3 Assignment 1 to calculate the following
    for each stratum:

    a.  mean grasshoppers per stratum (`y_bar`)

    ```{r}
    # take the mean of the `abund` column
    mean_A <- 
    mean_B <- 
    mean_C <- 
    ```

    b.  total grasshoppers per stratum (`tau_hat`)

    ```{r}
    tau_hat_A <- 
    tau_hat_B <- 
    tau_hat_C <- 
    ```

    c.  sample variance per stratum (`var` or s\^2)

    ```{r}
    # similar to (a) but calculating the variance instead of the mean
    var_A <- 
    var_B <- 
    var_C <- 
    ```

#### Step 2: Calculating Uncertainty (per stratum)

Our next step is to calculate the uncertainty in our estimates for each
stratum.

14. To do this, calculate the following:

    a.  variance of the estimate of the mean per stratum (var_y\_bar).
        Remember that we need to use the finite population correction
        factor and to use only objects already created, not plain
        numbers! (2 points)

    ```{r}
    pop_correction_A <- 
    pop_correction_B <- 
    pop_correction_C <- 

    var_y_bar_A <-
    var_y_bar_B <-
    var_y_bar_C <-
    ```

    b.  variance of the estimate of the total abundance (var_tau_hat).
        You might need to go fishing for this equation in one of the
        earlier powerpoints.

    ```{r}
    var_tau_hat_A <-
    var_tau_hat_B <-
    var_tau_hat_C <-
    ```

#### Step 3: Calculating Total Abundance Statistics

Congratulations! The hardest parts of the calculation are over (thank
goodness, right?).

Next, we will calculate the total abundance and the variance of the
total abundance.

15. Calculate the following:

    a.  total abundance

    ```{r}
    tau_hat <- 
    ```

    b.  variance of the total abundance estimate

    ```{r}
    var_tau_hat <- 
    ```

#### Step 4: Calculating Population Means, etc.

Finally, we can now calculate the overall population mean and variance
of the estimate of the overall population.

16. Calculate the following:

    a.  overall population mean

    ```{r}
    y_bar <- 
    ```

    b.  variance of the estimate of the overall population mean

    ```{r}
    var_y_bar <- 
    ```

#### Conclusion

Ok, that was a lot of calculating! Let's summarize and make sure we know
what just happened. We were ultimately able to calculate estimates the
parameters we were interested in: the population mean and the population
total along with the amount of uncertainty in both of those estimates.

17. Based on our calculations above, answer the following questions with
    the name of the R object and its corresponding value for each
    calculation. (4 points)

    a.  the number of grasshoppers per sample unit:

    b.  total number of grasshoppers in the population:

    c.  measure of uncertainty in the \# of grasshoppers per sample
        unit:

    d.  measure of uncertainty in the total \# of grasshoppers:
