---
title: "Module 2 Assignment 1"
author: "Ellen Bledsoe"
date: '2022-10-11'
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message = FALSE}
rocks <- read.csv("../data_raw/example_data.csv")
rocks
N <- 64
n <- 8
```

### Remember, these numbers will not match your data!

These values match the values in the `example_data.csv`. I've provided
them in case you want to check your code with the example data to make
sure you have done the correct thing.

If you do check your calculations using the `example_data.csv`:

- *Temporarily* replace the `rocks_abund.csv` file with the
  `example_data.csv` file in the `read.csv()` function in Question 5 and
  re-run the code chunk.
- Now, re-run all of the following code chunks. Doing so will compute
  all of the calculations with the "new" rocks data from
  `example_data.csv`.
- You can now compare your outputs to the outputs in this Answer Key. If
  you did all of the calculations correctly, the answers in your
  assignment should match the ones below.
- Once you are done, be sure to change the file in Question 5 back to
  your `rocks_abund.csv` file before you knit to PDF and submit your
  assignment!

5.  

```{r}
rocks_mean <- mean(rocks$abund_dark_stones)
rocks_mean
```

9.  

```{r}
rocks_var <- var(rocks$abund_dark_stones)
rocks_var
rocks_sd <- sd(rocks$abund_dark_stones)
rocks_sd
```

10. 

```{r}
pop_correction <- (N-n)/N
pop_correction
```

12. 

```{r}
rocks_var_ybar <- pop_correction * (rocks_var/n)
rocks_var_ybar
```

13. 

```{r}
rocks_SE <- sqrt(rocks_var_ybar)
rocks_SE
```

14. 

```{r}
rocks_total <- N * rocks_mean
rocks_total
```

15. 

```{r}
rocks_var_total <- N * (N-n) * (rocks_var/n)
rocks_var_total
```

16. 

```{r}
rocks_SE_total <- sqrt(rocks_var_total)
rocks_SE_total
```
