---
title: 'Module "4": Assignment 1'
author: "Ellen Bledsoe"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment Details

### Purpose

The goal of this assignment is to understand, apply, and interpret detection probability in estimating densities through distance sampling.

### Task

Write R code to successfully answer each question below or write text to successfully answer the question.

### Criteria for Success

-   Code is within the provided code chunks
-   Code chunks run without errors
-   Code produces the correct result
    -   Code attempts will get half points
    -   Code that produces the correct answer will receive full points
-   Text answers correctly address the question asked

### Due Date

May 6 at midnight MST

------------------------------------------------------------------------

# Assignment Questions

All questions are worth 1 point unless otherwise specified.

## Set-Up

We are going to need to load the community ecology package, `vegan`.

```{r message = FALSE}
library(vegan)
```

## Species Accumulation Curves

For this section of the assignment, we are going to use a survey by species matrix of data collected from the Santa Cruz River on trail cameras!

```{r}
scr_data <- read.csv("SRC_camera_data.csv")
```

1.  First, use the `specaccum()` function to prep the data for the species accumulation curve model.

```{r}

```

2.  Next, fit a species accumulation model to the data.

```{r}

```

3.  What is the predicted asymptote for this data?

```{r}

```

4.  Let's plot species accumulation curve model.

```{r}

```

5.  Based on the predicted asymptote, the number of species we have in the dataset, and the plot you just made, what can we conclude? (2 points)

*Answer:*

## Species Richness and Diversity Indices

Run the following code chunk to create 2 different data frames with rodent community data.

```{r}
comm1 <- data.frame(species = c("DO", "DM", "SO", "SH", "SF", "PP", "PB", "BA"),
                    individuals = c(14, 89, 30, 12, 3, 3, 8, 23))
comm2 <- data.frame(species = c("DO", "DM", "SO", "SH", "SF"),
                    individuals = c(37, 32, 30, 44, 39))
```

### Species Richness

6.  Use the `nrow` function to calculate the species richness of each community. Which community is more species rich?

```{r}

```

### Shannon Diversity Index

We are going to start with `comm1`.

7.  Write a line of code to calculate the number of individuals in each community. Use this to create the object `N_comm1`.

```{r}

```

8.  Add a column to the `comm1` data frame with each species' proportion of the community.

```{r}

```

9.  Now add a column with the natural logs of the proportions using the `log()` function.

```{r}

```

10. Add a column with the result of the proportion of the community multiplied by the natural log.

```{r}

```

11. Calculate the Shannon Index for this community.

```{r}

```

12. Now that you've worked your way through calculating the Shannon Index for community 1, calculate it for community 2.

```{r}

```

13. Which community is more diverse, according the the Shannon Diversity Index? How do you know? (2 points)

*Answer:*

### Simpson Diversity Index

Fortunately, a lot of the work has already been done to calculate the Simpson Diversity Index. We only need to add a little bit more to get there.

14. Add another column to the `comm1` data frame that has the squared proportion.

```{r}

```

15. Calculate the Simpson Diversity Index for community 1

```{r}

```

16. Now calculate the Simpson Index for community 2

```{r}

```

17. Do the values you've calculated for the Simpson Index corroborate your answer for which community is more diverse? Why or why not? (2 points)

*Answer:*
