---
title: "Honors Project Fall '23"
author: "Karen Ornelas"
output: 
  pdf_document: default
date: "2023-09-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will be using portalr for this project so first we must load in the package and tidyverse.

```{r}
install.packages("portalr")
library(portalr)
library(tidyverse)
```

```{r}
plant_data <- summarise_plant_data(path = "repo", level = "Quadrat")
```

Create an object where we we tell R to conditionally format our data into 3 specific criteria -\> Species = Amaranthus palmeri, Year = 2021, and Season = summer. This will serve as the population for the project.

```{r}
amar_2021 <- plant_data[
                        plant_data$species =="amar palm" 
                        & plant_data$year == 2021 
                        & plant_data$season == "summer", ]
```

```{r}
plot_number <- unique(plant_data$plot)

quadrat_number <- unique(plant_data$quadrat)

all_quads <- expand.grid(plot = plot_number,quadrat = quadrat_number)
```

Merge amar_2021 with all_quads data based on plot and quad

```{r}
expanded_amar <- dplyr::full_join(amar_2021,all_quads)
expanded_amar <- dplyr::mutate(expanded_amar, abundance = tidyr::replace_na(abundance, 0))
```

Calculations of our measures of central tendency

```{r}
mean <- mean(expanded_amar$abundance)
median(expanded_amar$abundance)  
```

Calculations of our measures of dispersion

```{r}
min(expanded_amar$abundance)
max(expanded_amar$abundance)
sd <- sd(expanded_amar$abundance)
var(expanded_amar$abundance)
```

We must first calculate our plot and quad area to find the N a single plot. Once this is found we can calculate by 24 (total number of plots).

```{r}
plot_length <- 50 
plot_area <- plot_length^2
quad_area <- 0.25
N_plot <- plot_area/quad_area
N <- 24 * N_plot
n <- nrow(expanded_amar)
```

```{r}
hist(expanded_amar$abundance)
```

```{r}
### install tidyverse
install.packages("tidyverse")
library(tidyverse)
```

Calculations of the occupancy

```{r}
amar_occu <- expanded_amar[ , c("plot", "quadrat", "abundance")]
```

Lets change anything in amar_occu above a 0 to 1. This makes are data either a 0 or 1 ( absent or present.)

```{r}
### change anything above 0 into a 1
amar_occu$abundance <- ifelse(amar_occu$abundance >0, 1, 0)
```

This will turn our format from plot,quad, abund to plot and quad as our columns and abundance as our rows.

```{r}
amar_occu <- pivot_wider(amar_occu, names_from = quadrat, values_from = abundance )

###remove pilot
amar_occu <- amar_occu[, -1]
```

```{r}
install.packages("unmarked")
library(unmarked)
```

Now we must create a data frame to have a list of our surveys.

```{r}
surveys <-data_frame(time_1 = rep("1", 24), time_2 = rep("2", 24), time_3 = rep("3", 24), time_4 = rep("4", 24), time_5 = rep("5", 24), time_6 = rep("6", 24), time_7 = rep("7", 24), time_8 = rep("8", 24), time_9 = rep("9", 24), time_10 = rep("10", 24), time_11 = rep("11", 24), time_12 = rep("12", 24), time_13 = rep("13", 24), time_14 = rep("14", 24), time_15 = rep("15", 24), time_16 = rep("16", 24) )
```

```{r}
### create list of dataframe

surveys <- list(time = surveys)

```

Create an unmarked data frame that will be used to calculate our occupancy models.

```{r}
amar_umf <- unmarkedFrameOccu(amar_occu, obsCovs = (surveys))
```

p(.) psi(.) -\> occupancy model where occupancy and detection probability are constant p(time) psi(.) -\> occupancy model where occupancy stays constant but detection probability varies with survey.

```{r}
amar_pdot_psidot <- occu(~1 ~1, data = amar_umf)

amar_ptime_psidot <- occu(~time-1 ~1, data = amar_umf)
```

```{r}
amar_models <- fitList(amar_pdot_psidot = amar_pdot_psidot , amar_ptime_psidot = amar_ptime_psidot)

modSel(amar_models)
```

```{r}
#Dectection Prob
plogis(coef(amar_pdot_psidot))
```

```{r}
#CI for detection
plogis(confint(amar_pdot_psidot, type = "det", level = 0.95))
```

```{r}
#Prob Occu
plogis(coef(amar_pdot_psidot))
```

```{r}
#CI for occupancy
plogis(confint(amar_pdot_psidot, type = "state", level = 0.95))
```

The AIC shows that p(.) psi(.) would be the best model because of the lower AIC model. 10% detection probability with a 95% CI range of 6% - 15%. Occupancy level of 80% with a 95% CI range of 43-95%. The detection and occupancy differences can be due to environmental factors such as rainfall, temperature, and plant cover.

```{r}
### md 2 
Tau_hat <- N * mean
Var_ybar <- (N-n)/N * (sd^2 / n)
Var_tau_hat <- N * Var_ybar
SE_ybar <-  sqrt(Var_ybar)
SE_tau_hat <- sqrt(Var_tau_hat)
K <- N/n
```

```{r}
Tau_hat
Var_ybar
Var_tau_hat
SE_ybar
SE_tau_hat
K
```

This is a 1 in 625 systematic sample.

The number of amaranth per sample unit: y_bar: 0.18 Total number of amaranth in the population: tau-hat: 43750 Measure of uncertainty in the \# of amaranth per sample unit: var_y_bar: 0.007 Measure of uncertainty in the total number of amaranth: var_tau_hat: 1738.79
