---
title: "Extra Credit"
author: "Your Name Here"
output: 
  pdf_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Extra Credit Project

For this EC project, you will be calculating various population paramters and their uncertainty based on what we have learned throughout RNR 321.

You will use two different types of data from the [Portal Project](https://portal.weecology.org/). It is worth reading at least a little bit about the project to help contextualize what you will be doing.

## Set Up

We will be using the `portalr` and `tidyverse` packages for this project, so first we must load them both.

```{r}
install.packages("portalr")
install.packages("tidyverse")

library(portalr)
library(tidyverse)
```

We will be calculating population metrics and running occupancy models for *Amaranth palmerii* ("amar palm").

First, we have to download the data.

```{r}
plant_data <- summarise_plant_data(path = "repo", level = "Quadrat")
```

Create an object with only "amar palm" plants from the summer plant census of 2021. This will serve as the population for the project.

```{r}
amar_2021 <- plant_data %>% 
  filter(species == "amar palm", 
         year == 2021,
         season == "summer")
```

Make an expanded grid to include every quadrat at every plot so we include zeros in our estimate.

Plant surveys occur at 16 different locations in each of the 24 plots. Each survey includes the species present and their abundance in a 0.5 m by 0.5 m quadrat (0.25 m\^2).

```{r}
# find unique plots and unique stakes
plot_number <- sort(unique(plant_data$plot))  # 24 plots
quadrat_number <- sort(unique(plant_data$quadrat))  # 16 quadrats per plot

# create a dataframe with every quadrat in every plot
all_quads <- expand.grid(quadrat = quadrat_number,
                         plot = plot_number)


```

Next, we need to bring together the `amar_2021` data with `all_quads` data and fill empty values with 0 to indicate absence.

```{r}
treatments <- plant_data %>% 
  filter(year == 2021) %>% 
  select(plot, treatment) %>% 
  distinct()

amar_2021 <- full_join(amar_2021, all_quads) %>% 
  # replace NA values with 0
  mutate(abundance = replace_na(abundance, 0),
         treatment = case_when(plot %in% treatments$plot[treatments$treatment == "removal"] ~ "removal",
                               plot %in% treatments$plot[treatments$treatment == "exclosure"] ~ "exclosure",
                               TRUE ~ "control")) %>% 
  select(quadrat, plot, abundance, treatment)
```

## Module 1

Calculate measures of central tendency (median and average). Save these outputs to use in future calculations.

```{r}
mean <- mean(expanded_amar$abundance)
median(expanded_amar$abundance)  
```

Calculate measures of dispersion (range, standard deviation, variance). Save these outputs to use in future calculations.

```{r}
min(expanded_amar$abundance)
max(expanded_amar$abundance)
sd <- sd(expanded_amar$abundance)
var(expanded_amar$abundance)
```

Plot of histogram of the "amar palm" abundance (it will look very skewed; that's ok).

```{r}
hist(amar_2021$abundance)
```

## Module 2

We need to figure out the number of sample units per plot and then multiple that number by 24 to get the total number of sample units for the site.

```{r}
# each plot is 50 x 50 m
plot_area <- 50^2
quad_area <- 0.25

plot <- plot_area/quad_area

N <- 24 * plot
n <- nrow(amar_2021)
```

### Stratified Random Sample Calculations

The Portal Project has 3 different treatments; each plot falls into 1 treatment group:

-   removal = no rodents in the plot
-   exclosure = no kangaroo rats on the plots
-   control = all rodents on the plots

Because rodents are herbivores, we have reason to believe they could impact the plant communities on different plot types.

Let's treat the data as a stratified random sample, with the treatment type as the stratum.

Make 3 new dataframes, one for each stratum based on plot treatment, by subsetting the data using conditional subsetting (base R or `tidyverse`).

```{r}
removal <- amar_2021 %>% filter(treatment == "removal")
exclosure <- amar_2021 %>% filter(treatment == "exclosure")
control <- amar_2021 %>%  filter(treatment == "control")
```

Use the `nrow()` function and the dataframes above to determine the `n` (number of sample units) for each stratum.

```{r}
n_removal <- nrow(removal)
n_exclosure <- nrow(exclosure)
n_control <- nrow(control)
```

Here are the `N` values for each stratum:

```{r}
N_removal <- N * (sum(treatments$treatment == "removal") / 24)
N_exclosure <- N * (sum(treatments$treatment == "exclosure") / 24)
N_control <- N * (sum(treatments$treatment == "control") / 24)
```

Now, using these `n` and `N` values, and the dataframes above, calculate the following values:

-   stratum mean
-   stratum variance
-   stratum variance of the estimate of the mean
-   stratum total
-   stratum variance of the estimate of the total

Remember to include the finite population correction factor where appropriate!

```{r}

```

### Overall Population Estimates and Uncertainty

Using the appropriate equations and the values above, calculate the following values for the entire population:

-   overall mean
-   overall variance of the estimate of the mean
-   overall total
-   overall variance of the estimate of the total

```{r}

```

Assuming a t-value of 2.06, calculate the standard error and confidence intervals for the overall mean and population total.

```{r}

```

Write a few sentences interpreting your values. For example: (a) which treatment has the highest/lowest density of amaranth? (b) which treatment's estimates have the most uncertainty around them? (c) what is the overall population density and abundance of amaranth at Portal?

*Answer*:

## Module 3

For this module, we are going to calculate occupancy of "amar palm." We are going to ignore treatments here and consider the population as a whole.

We need to load `unmarked`.

```{r}
library(unmarked)
```

We are going to do this in a somewhat unique way. We are going to consider each plot as a survey location and each quadrat as a separate survey.

The following code chunk converts the data into an encounter history that we can use.

```{r}
amar_occu <- amar_2021 %>% 
  mutate(abundance = if_else(abundance > 0, 1, 0)) %>% 
  pivot_wider(amar_occu, names_from = quadrat, values_from = abundance) %>% 
  select(-plot)
```

Here, we create our surveys dataframe so we can compare models where detection probability varies by survey (quadrat) or not.

```{r}
surveys <- data_frame(time_1 = rep("1", 24), time_2 = rep("2", 24), time_3 = rep("3", 24), time_4 = rep("4", 24),
                      time_5 = rep("5", 24), time_6 = rep("6", 24), time_7 = rep("7", 24), time_8 = rep("8", 24), 
                      time_9 = rep("9", 24), time_10 = rep("10", 24), time_11 = rep("11", 24), time_12 = rep("12", 24), 
                      time_13 = rep("13", 24), time_14 = rep("14", 24), time_15 = rep("15", 24), time_16 = rep("16", 24))
surveys <- list(time = surveys)
```

Follow the steps in Module 3 Lab 1 to run the occupancy models for "amar palm."

```{r}

```

Determine which model fits the data best based on AIC.

```{r}

```

Calculate and report (a) occupancy, (b) detection probability, and (c) the confidence intervals for each

```{r}

```

Write a few sentences interpreting the results.

*Answer*:

### Module 4

We will keep this simple. Calculate species richness for the summer plant survey in 2021. Using the `plant_data` data frame, filter the data appropriately and then use the `unique()` function.

```{r}

```
